%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%   NSF RTG  2023
%   Major revision starting March 2023. 
% PIs: Sonja, Chun, Fred, Xiaofan, and ?? 
% senior personnel: Hemanshu, Rob, Michael, Despina (res&SoReMo+math), Yuhan(for summer UG), Gorjana (editorial/student training), Trevor Lesile, Yubin Lu, WILL ADD OTHERS SOON 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%   NSF RTG  2023
%   Revising for 2023 submission! 
%   Due:   June 6, 2023
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[11pt]{NSFamsart}

%%%%%%%%%%%%%%% Margins and Page Numbers %%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\voffset 0.23in  %Made to satisfy Research.gov compliance
\hoffset 0.01in
\textheight 8.91in
\textwidth 6.46in
\setlength{\oddsidemargin}{0in}
\setlength{\evensidemargin}{0in}
\setlength\marginparwidth{60pt}

%\thispagestyle{empty} \pagestyle{empty} %to eliminate page numbers for upload
\thispagestyle{plain} \pagestyle{plain} %to add back page numbers

%%%%%% Squeeze the space %%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
%\renewcommand{\baselinestretch}{0.97}
%%%%%%Squeeze the space %%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 

\headsep-0.6in

%%%%%%%%%%%%%%% Margins and Page Numbers %%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 

%\oddsidemargin 0.10in \evensidemargin -0.65in
%\textwidth 6.2in         % Width of text line.
%\topmargin 0.60in \headheight 0.0in \headsep 0.0in
%\textheight 8.5in        % Height of text (including footnotes and figures,
%\topskip 0.0in

%  \usepackage{showkeys}
\usepackage{xcolor}
\usepackage[numbers,sort&compress]{natbib}
\usepackage{amsmath, amssymb, latexsym, xspace}
\usepackage{psfrag,epsfig,amsfonts,amsmath,latexsym,amsthm,amssymb,amscd,url }
\usepackage{amsmath}
\usepackage{bm}
\usepackage{enumitem,verbatim}
\usepackage[normalem]{ulem} 


\setlist[description]{font=\normalfont\bfseries, labelindent = 0cm, leftmargin=0ex}
\setlist[itemize]{leftmargin=5ex}


\newcommand{\F}{{\mathcal{F}}}
\newcommand{\B}{{\mathcal{B}}}
\providecommand{\HickernellFJ}{Hickernell}

\newcommand{\eps}{\varepsilon} 

\renewcommand{\k}{\kappa}
\newcommand{\p}{\partial}
\newcommand{\D}{\Delta}
\newcommand{\om}{\omega}
\newcommand{\Om}{\Omega}
\renewcommand{\phi}{\varphi}
\newcommand{\e}{\epsilon}
\renewcommand{\a}{\alpha}
\renewcommand{\b}{\beta}
\newcommand{\N}{{\mathbb N}}
\newcommand{\R}{{\mathbb R}}
\newcommand{\T}{{\mathbb T}}

\newcommand{\Le}{L_t^{\alpha}}

\newcommand{\EX}{\mathbb{E}}
\newcommand{\PX}{\mathbb{P}}


\newcommand{\grad}{\nabla}
\newcommand{\n}{\nabla}
\newcommand{\curl}{\nabla \times}
\newcommand{\dive}{\nabla \cdot}

\newcommand{\ddt}{\frac{d}{dt}}
\newcommand{\la}{{\lambda}}

\newcommand{\bu}{\mathbf{u}}

\newcommand{\obu}{\bar{\mathbf{u}}}
\newcommand{\bsigma}{\mathbf{\sigma}}
\newcommand{\btau}{\mathbf{\tau}}


\newcommand{\nd}{{\nabla \cdot}}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\x}{{\bm x}}

\newcommand{\cF}{{\cal F}}
\newcommand{\cG}{{\cal G}}
\newcommand{\cD}{{\cal D}}
\newcommand{\cO}{{\cal O}}

\newcommand{\numUG}{33\xspace}
\newcommand{\numMS}{8\xspace}
\newcommand{\numPHD}{25\xspace}
\newcommand{\numGrad}{33\xspace}
\newcommand{\numPostDoc}{5\xspace}
\newcommand{\numjoint}{94\xspace}

%%%%%%%%%%%%%%

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{definition}{Definition}
 \newtheorem{coro}[lemma]{Corollary}
 \newtheorem{example}[lemma]{Example}
 \newtheorem{remark}[lemma]{Remark}


%%%%%%%%%%%%%%% %%%%%%%%%%%%%



\newcommand{\uk}[1]{\ensuremath{u^{(#1)}(t,\omega)}}
\newcommand{\hse}{\ensuremath{h^s(\xi,\omega)}}
\newcommand{\hsk}[1]{\ensuremath{h^{(#1)}(\xi,\omega)}}

\newcommand{\sz}{\ensuremath{ {\int_s^0 z(\theta_r (\omega))\,dr}}}
\newcommand{\sZ}{\ensuremath{ {\int_s^0 Z(\theta_r (\omega))\,dr}}}
\newcommand{\zz}[1]{\ensuremath{{z(\theta_{#1} (\omega))}}}
\newcommand{\ZZ}[1]{\ensuremath{{Z(\theta_{#1} (\omega))}}}
\newcommand{\fu}[1]{\ensuremath{{F_u^{u_0 (#1)}}}}
\newcommand{\fus}[2]{\ensuremath{{\int^0_{#2} F_u^{u_0 (#1)}\,d{#1}}}}
\newcommand{\fuss}[2]{\ensuremath{{\int^{#2}_0 F_u^{u_0 (#1)}\,d{#1}}}}

\newcommand{\fuu}[1]{\ensuremath{{F_{uu}^{u_0(#1)}}}}
\newcommand{\rb}{\right)}
\newcommand{\lb}{\left(}
\newcommand{\rB}{\right]}
\newcommand{\lB}{\left[}


\newcommand{\nb}{\mathbf{n}}
\newcommand{\ub}{\mathbf{u}}
\newcommand{\xb}{\mathbf{x}}
\newcommand{\xnb}{\mathbf{x}_0}
\newcommand{\GaB}{\mathbf{\Gamma}}

\newcommand{\bo}{\mathcal {O}}
\newcommand{\so}{\mathcal {o}}

\newcommand{\BE}{\begin{equation}}
\newcommand{\EE}{\end{equation}}
\newcommand{\BEN}{\begin{equation*}}
\newcommand{\EEN}{\end{equation*}}
\newcommand{\BAL}{\begin{align}}
\newcommand{\EAL}{\end{align}}
\newcommand{\BAN}{\begin{align*}}
\newcommand{\EAN}

%\newcommand{\s}{\sigma}
\def\Tr{\mbox{Tr}}
\newcommand{\Rn}{\mathbb{R}^n}

\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}
\newcommand{\FredNote}[1]{{\color{blue} Fred: #1}}
\newcommand{\YuhanNote}[1]{{\color{violet} Yuhan: #1}}
\newcommand{\bfparagraph}[1]{\paragraph{\textbf{\textup{#1}}}}

\iffalse
The long-range goal of the Research Training Groups in the Mathematical Sciences (RTG) program is to strengthen the nation's scientific competitiveness by increasing the number of well-prepared U.S. citizens, nationals, and permanent residents who pursue careers in the mathematical sciences, be they in academia, government, or industry. A significant part of this goal is to directly increase the proportion and the absolute number of U.S. students at the RTG sites who pursue graduate studies and complete advanced degrees in the mathematical sciences. It is anticipated that RTG projects also will serve as national models for research training in the mathematical sciences. Activities with potential impact beyond the directly-supported students and beyond the institutions receiving RTG funds will be key strengths in proposals. Collaborative proposals involving different types of programs (for example, institutions in which the relevant department does not award Ph.D.s, minority-serving institutions, etc.) and having the potential to develop innovative approaches to research training in the mathematical sciences are welcome. For such collaborative efforts, the lead institution must grant a doctoral degree in mathematical sciences.

%% **  SONJA READ THIS AGAIN ** 
The RTG program supports efforts to improve research training by involving undergraduate students, graduate students, postdoctoral associates, and faculty members in structured research groups anchored in a coherent research program. The activities need not be focused on a particular research problem; rather, it is expected that group participants will be united by common topical interests. The groups may include researchers and students from different departments and institutions, but the research-based training and education activities must be based in the mathematical sciences. RTG projects are expected to vary in size, scope, and proposed activities, as well as in their plans for organization, participation, and operation. However, research groups supported by RTG will include vertically-integrated activities that span the entire spectrum of educational levels from undergraduates through postdoctoral associates.
Addressing all stages (from undergraduate through postdoctoral) of trainee involvement is essential in RTG proposals. Proposals that focus on only one stage are not appropriate for submission to the RTG activity. While emphasis on graduate training in RTG projects is appropriate and natural, a substantial plan for involving undergraduates is necessary. When used in reference to undergraduates, the word "research" should be given its broadest interpretation.

%% **  SONJA READ THIS AGAIN ** 
Successful proposals will include collaborating faculty with a history of research accomplishments. This group should have a history of working with students and/or postdoctoral associates, and they should present a strong plan for recruiting students who are U.S. citizens, nationals, or permanent residents into their program. The RTG program is not meant to establish new research groups, but to enhance the training activities of existing groups with strong research records.

Graduate Trainee-ships. Graduate trainees form a pivotal component of the integration of activities in RTG grants. Their participation should result in:
1. involvement with research activities that include undergraduates, other graduate students, postdoctoral associates, and/or faculty members; 
2. graduate education that is both broad and deep; and
3. significant teaching or other professional experience such as industry/laboratory internship.
Mentoring, that is, guidance in professional development, is a critical strategy for preparing graduate trainees to become successful researchers, communicators, and mentors. Graduate trainees are expected to have substantial mentored professional experiences to prepare them for successful careers in the mathematical sciences and in other professions in which expertise in the mathematical sciences plays an important role. Examples of this professional experience could include:
• a minimum of two terms of supervised teaching, preferably with one term of more independent teaching in which the student has substantial responsibility for a class, or
• a minimum of two terms of a supervised industry/laboratory internship.
Some element of their activities should help students develop proficiency in the presentation of mathematical sciences research in both written and oral formats and in the ability to place their research in context.

RTG awards are intended to allow graduate students significant time for research, course work, and related activities. A graduate trainee can receive up to 33 months of non-teaching support from an RTG activity. RTG stipends cannot be used to pay students to fulfill teaching duties or for internships. Departments must demonstrate how the traineeships will improve the quality of the education their graduate students receive. The traineeships are not intended to replace existing institutional funding of research fellowships or scholarships.

%% **  SONJA READ THIS AGAIN ** 
Undergraduate Experience. In this program solicitation, the term "research experiences" for undergraduates includes all activities that involve undergraduates in discovery and generate appreciation of and excitement about research in the mathematical sciences. An undergraduate research experience does not have to result in the publication of a paper. Examples of research experiences include faculty-directed projects, either during the academic year or the summer, or participation in research teams with graduate students and/or postdoctoral associates. Such experiences are intended to involve students in the creative aspects of mathematical sciences in a non-classroom setting. They are also expected to enhance the development of students' communication skills, with particular emphasis on the presentation of mathematical concepts in both written and oral formats. In all cases, it is expected that the participating undergraduates receive mentoring to stimulate their further interest in the mathematical sciences.

Postdoctoral Associates. Effective RTG activities better prepare postdoctoral associates for their future careers. It is expected that at the end of the postdoctoral experience, each associate will have a well-defined independent research program, well-developed communication skills, a broad perspective of his or her field, and the ability to mentor.

The postdoctoral program can provide opportunities not traditionally found in mathematical sciences education and training, including interdisciplinary research experiences in connection with other departments and programs; participation in international research programs; internships in business, industry, or government laboratories; or participation in research institute programs suitably aligned with the associate's research interests. Postdoctoral associates are expected to teach, on average, one course per term while in residence at the sponsoring university. Over the duration of the postdoctoral appointment, this teaching should encompass a diverse set of instructional experiences at different levels of the curriculum. Likewise, it is expected that each RTG postdoctoral associate will submit a research proposal to a funding agency at some time during the course of the postdoctoral appointment. Mentoring to help ensure all postdoctoral associates become successful researchers, communicators, and mentors is a critical element of an RTG postdoctoral program, as is interaction of postdoctoral associates with undergraduate and/or graduate students.

The typical RTG postdoctoral appointment is for three years. A person is eligible for only one RTG postdoctoral appointment. An RTG postdoctoral associate is expected to be a recent recipient of a doctoral degree, typically held not more than three years as of January 1 of the year in which the appointment begins. Any exceptions made to this restriction should be well-justified in the annual reports.

Budget. Proposals may include support requests for graduate and advanced undergraduate students, postdoctoral associates, visitors, consultant services, travel, conferences, and workshops. Other budget items that are deemed to be essential to the success of the proposed activities may be included. Faculty salary is limited to that needed for the purpose of organizing and managing the program.

Data. RTG proposals will be strengthened by supporting data about the department's programs. An extensive discussion of the requested data appears in the Supplementary Documentation section below (V.A.7).

A successful RTG proposal will:
• be based in a U.S. IHE that grants the Ph.D. in the mathematical sciences (faculty and trainees from other types of institutions may be included through a collaborative proposal or other mechanisms);
• be anchored in a coherent research program in the mathematical sciences;
• have a realistic plan showing how the proposed activity would create new or enhanced research-based training experiences in the mathematical sciences for the students and postdoctoral associates;
• be directed by a principal investigator, with at least two other faculty members, who will collaborate in management and participate fully in the RTG activities.
A successful RTG proposal must convince reviewers that the project:
• integrates research with educational activities;
• provides for developing professional and personal skills, such as communication, teamwork, teaching, mentoring, and leadership;
• includes an administrative plan and organizational structure that ensures effective management of the project resources;
• has an institutional commitment to furthering the plans and goals of the RTG project and to create a supportive environment for integrative research and education;
• has a plan for recruitment, selection, and retention of participants, including members of underrepresented groups, so as to increase the number and diversity of U.S. citizens, nationals, and permanent residents in the graduate and postdoctoral programs;
• serves as a national model by effectively disseminating best practices for attraction, retention, and high-quality preparation of students and postdoctoral associates in the mathematical sciences; and
• has a post-RTG plan. The RTG program is intended to help stimulate and implement permanent positive changes in research training within the mathematical sciences in the U.S. Thus it is critical that an RTG site adequately plan how to continue the pursuit of RTG goals when funding terminates.
\fi

\iffalse
Guided by data, market intelligence, and trends, we are focusing on high growth sectors, accelerated by technology and innovation, from start-up to Fortune 500.

The World Business Chicago Research Center curates original content and data analysis on many facets of the city’s diverse economy. A monthly publication, Chicago Business Bulletin, digs into specific sectors realizing significant growth to identify how Chicago’s inherent strengths and attributes, coupled with tech and innovation, are driving expansion and growth.

As rapid expansion of computation and data creates seismic shifts in business and industry, the College of Computing equips its graduates to capitalize on this phenomenon to drive innovation and decision-making.

Illinois Tech founded the College of Computing in 2020 to ensure student success in the fast-growth fields of computation and data sciences. Our location in the burgeoning tech center of Chicago provides students with access to meaningful internship and career opportunities found only in a world-class city. As a member of the Discovery Partners Institute, Illinois Tech is a gateway to a wider world of research and development. Connections with P33 and other entities in Chicago extend career pipelines to the city’s leading tech businesses.

The College of Computing develops the talent, tools, and technology to fuel a growing tech industry in Chicago and beyond by educating a workforce sophisticated in cutting-edge fields of computer science, applied mathematics, and information and industrial technologies. Our breadth gives us expertise and instruction from the very foundational ideas in computing to the very practical as well as industrial applications. Our renowned faculty are creating new boundaries in these fields through world-class research and education, offering both undergraduate and graduate students opportunities to play a role in discovering significant breakthroughs and solving real-world problems.
\fi




%%%%% Proposal Starts Here:     %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}  

\centerline{\large \textbf{RTG: Sampling, inference,  dynamics, and their interactions in data science}}
%for inference .... %inference, and dynamics for  ... data /applications }}
%Research and Training in Complex Dynamical Systems}}

\begin{comment}
\section*{\color{magenta}Action items list:}

{\color{magenta}BUDGET NOTES: \\
* 5 postdocs (hire 2 in 1st year, 1 in 2nd year, 2 in 3rd year); \\
* 1-2 weeks summer salary for one PI per year - Chun in year1, others rotate after;\\
* PhD: 2 per year = total 10;\\
* undergrads: 6 per year -- this includes SoReMo fellows FROM MATH (AY UG); plus SURE (SUmmer UG); \\
* Rob's I-course funding for students - TA grad student training for this course (applied data science!);\\
* Despina's training module for student research project proposals --about 10K per year; \\
* ... 
}
* \sout{update title of the proposal LOL} 

* \sout{remove Duan from the proposal}

* \sout{add Sonja's research section draft}

* \sout{rewrite and polish Sonja's research section draft.  Work-in the joint work with the discrete group.} Polish Thrust 2 Sonja-Hemanshu-Micheal part.

* \sout{*Sonja will also add Rob in the context of I-course (grad student training for this course (applied data science!)); Despina and Gorjana to SoReMo and training* Despina's training module for student research project proposals } --about 10K per year?? this is coming out as summer salary for SPs;

* Yuhan will be senior personnel in the context of SURE. 


* \sout{[THIS GOT LOST DURING THE SUMMER] Sonja, see line 171 \& 175 for code comments from the solicitation!!  and 189.}

* Emphasize that our proposal includes senior, junior and NTT (teaching) faculty. 

* Add KI's enterpreneurship collaborations for students and open to postdocs - which is something unique to IIT, I think. 

* the rest of the text all needs updates, both quant and qual. :D 

* Introduction will be edited LAST. 
\end{comment}

\section{Introduction}

%\marginpar{\color{magenta}I worry about this first paragraph in general. We do not substantiate that we are going to attract Black or Latina students - and we do not have a track record of doing that? or we pretend we do, we have 1 Black 1 Latina mentioned in the proposal. I don't know, this does not sound convincing to me. :( It's the first thing we say, and it is not the focus of the RTG. }
The majority of Chicago's population are Black or Latine, 
%See https://www.dictionary.com/e/latine-vs-latinx/ for why Latine is better (it mirror's Spanish language
but they hold only 14\% of the best tech jobs \cite{P33}. 
Chicago strives to accelerate  technology and innovation in the city and beyond, guided by data, market intelligence, and trends \cite{Chicago}. Illinois Institute of Technology (Illinois Tech) is Chicago's technology-focused research university.  Our Department of Applied Mathematics---situated in the College of Computing---ensures student success in the fast-growing fields of mathematical modeling and simulation, prediction and decision-making, computation, and data science. 
The proposed Research Training Group (RTG) will grow a sophisticated and diverse talent pool for Chicago and the nation. 

% Sonja making somewhat big edits here; hope you agree it's a better summary!! 
Rooted in theoretical mathematical and statistical foundations, our focus is on strengthening our interdisciplinary academic program to train the next generation of applied mathematicians.  They will learn to creatively integrate tools from various areas to solve complex problems in technology, environment, finance, social and biomedical sciences. This RTG will establish  a reproducible and sustainable process for developing this needed maturity in mathematics graduate and undergraduate students and postdoctoral fellows. We will accomplish this with the following plan:  we will restructure student-centered research activities in the department (Section~\ref{section:regular research activities}), offer new opportunities and connect to existing summer activities in-house and with our partners (Section~\ref{sec:summer}), and institutionalize existing and develop new interdisciplinary co-curricular training activities (Section~\ref{section: co-curricular}). 

\begin{comment} 
While standard undergraduate and graduate mathematics curricula provide students with  analytical, computational, and statistical tools, these curricula generally do not help students \emph{mature} to the point that they can creatively integrate these various  tools to solve complex problems in technology, environment, finance, and biomedical science. This RTG will establish a reproducible and sustainable process for developing this needed maturity in mathematics graduate and undergraduate students and postdoctoral fellows.{\color{magenta}currently we do not substantiate these claims. return to this later.} We will accomplish this by developing new cross-disciplinary courses, updating existing courses, and integrating research experiences that cross disciplinary and institutional boundaries, and providing co-curricular activities that develop essential soft skills.  \FredNote{We need to hint what will be subtracted.  We cannot just add more.}We will recruit female and underrepresented minority students into our RTG through personal connections {\color{red}!!!} and our partner institutions.

% THIS IS NOT USEFUL AND IT JUST TAKES UP SPACE; I WROTE AN ALTERNATIVE PARAGRAPH BELOW.  -- SONJA. 
The  foundation???? of our training will be random sampling, statistical inference, and dynamics, and how mathematical foundations for these elements interact and are applied in data science. 
%\emph{complex dynamical systems}.  These dynamical systems may be subject to random influences \cite{Arnold, DuanBook2015}, far from equilibrium \cite{liu2009introduction}, or operate in quantum regimes \cite{Dittrich2016}. Our approaches for understanding them may be particle based (trajectory and sample path)  or continua based (probability density, ensemble, energy, action functional, and wave).  Our training in leading edge mathematical and computational tools will introduce students to a spectrum of application areas that are fundamental to current scientific and technological progress, such as geophysical and climate systems, liquid crystals, multi-phase flows, ionic fluids,  transportation systems, sensor networks, financial market risk, and biochemical reaction systems.
Sampling is used in .... {\color{magenta}pull in 2 of 3 Thrusts}, Inference plays out in ....  Dynamical systems may be subject to random influences \cite{Arnold, DuanBook2015}, far from equilibrium \cite{liu2009introduction}, or operate in quantum regimes \cite{Dittrich2016}.  
\end{comment} 


Scientifically, the RTG is centered around 3 {\bf research themes}: inference, dynamics, and sampling. Our research activities are grouped in 3 {\bf thrusts} described in Section~\ref{sec:researchproblems}, such that each thrust involves two of the research themes: 
%We propose three research thrusts: 
Thrust 1, \emph{Variational Analysis and Inference of High Dimensional Complex Dynamics}, couples inference and dynamics. 
%\marginpar{\color{cyan}TAG: MING, I don't know the final title of your thrust, edit as needed}
Thrust 2, \emph{Sampling combinatorial structures}, couples inference and sampling. 
Thrust 3, \emph{Dynamics under uncertainty}, couples sampling and dynamics. 
%All 3 s incorporate two of the main trusts of this RTG: namely, thrust 1 couples inference and dynamics, thrust 2 inference and sampling, and thrust 3 sampling and dynamics. 
Joint \FredNote{with what? across themes? or thrusts?}applications in data science provide an additional common thread.  %to the proposal. 
%under the umbrella of complex dynamical systems {\color{magenta}EDIT ME} NEW: thrust 1 = inference and dynamics; thrust 2 = sampling and inference; thrust 3 = sampling and dynamics. All 3 thrusts have interactions and data science connection. {\color{cyan}←\leftarrow This is the new THREAD of our proposal. } 1.\ Dynamics Under Uncertainty; 2.\  Energetic Variational Approaches for Complex Dynamics; and 3.\ Mathematical Tools for Open Quantum Systems. thrust 1 connects with Theme 2 via minimizing action functionals. In Thrust 3, we examine the interplay of quantum uncertainty and environmental fluctuations, and further consider a Bohmian trajectory formulation of quantum dynamics, thus linking with Thrust 1. 



Students and postdocs trained under this RTG will master theoretical and computational concepts that are the foundation of the three research thrusts. Through research seminars, projects, co-curricular training activities, and partnerships described in this proposal, they will be exposed to a broad spectrum of applications including  those fundamental to current scientific and technological progress: geophysical and climate systems, liquid crystals, multi-phase flows, ionic fluids,  transportation systems, sensor networks, financial market risk, social networks, socially responsible economic and infrastructure modeling, and biochemical reaction systems.

\subsection*{Academic Environment} The PIs are five Illinois Tech Applied Mathematics faculty:  %Jinqiao Duan\marginpar{remove Duan from the proposal start-finish; read through.}, 
Fred Hickernell, Xiaofan Li, Chun Liu, Sonja Petrovi\'c, and Ming Zhong. This RTG builds on the PIs' research expertise and our Illinois Tech applied mathematics curriculum.   The research portfolio of our department covers a broad range of applied mathematics and statistics, and includes topics germane to this RTG, including dynamical systems, stochastic dynamical systems, multiscale and variational methods,  Monte Carlo methods, foundations of computational mathematics, discrete mathematics, computational algebra, and data-driven predictive modeling and simulation. Our work has applications in the physical, chemical, geophysical, biophysical, and materials sciences, finance, and social sciences. 
% The PIs include present and past departmental, college, and university leadership.   Thus, we expect to persuade the bulk of our department to participate in this RTG in some capacity, whether as part of the research groups or implementing our curricular and co-curricular innovations.{\color{magenta}umm, this is actually *unethical*. Instead, we will *involve* a large portion of our colleagues as senior personnel and give them meaningful roles in this RTG, thereby contributing to its success and with a long-term training vision.}
The PIs and Senior Personnel (SPs) encompass every department group: senior and junior tenured and tenure track (TT) faculty and non-tenured/tenure track (NTT) teaching faculty. The involvement of a large portion of the department embeds this RTG into our daily activities in a way that contributes to both short-term success of the proposed program and a long-term transformative training vision. 

\marginpar{\color{cyan}TAG: Chun, can someone double-check these numbers? {\color{red}Yes, Chun will do }\color{cyan} remove CS students?? add DOMESTIC math students?}{\color{blue} we double checked the numbers} We currently have \numUG BS applied mathematics and statistics majors, \numGrad graduate students including \numMS master students and \numPHD doctoral students, \numjoint graduate students in programs offered jointly with the Department of Computer Science and \numPostDoc postdoctoral fellows or visiting assistant professors. Our graduates build their careers in  industry  and academia. Many of our undergraduates pursue PhDs at other top universities, including Cornell, NYU, UCLA and UIUC. Most of our postdoctoral researchers land in tenure-track positions after leaving Illinois Tech.

\subsection*{Partnership with  national labs and local institutes} 
%...{Argonne, IMSI, C2ST, and Oak Ridge}
This RTG will benefit from our ongoing collaboration with Argonne National Laboratory (18 miles west of our campus).  Argonne offers a variety of internship opportunities for undergraduate and graduate students, as well as faculty-student teams. Summer internships, supervised by Argonne staff, are typically 10-14 weeks long and involve training in computational mathematics tools and techniques. %Students work on current Argonne projects, supervised by Argonne staff. % The technical part of the internship is complemented by additional training opportunities.

Since 2007 the Chicago Council on Science and Technology (C2ST) has enhanced the public’s understanding and appreciation of science and technology and their impact on society. C2ST will co-sponsor outreach programs with our RTG and will train our students in communicating their scientific ideas to a broader audience.   

A recent joint research group was formed with peoples from various departments (ECE, CS and Aerospace) in Illinois Tech and the Data Analysis and Machine Learning Group within the Computer Science and Mathematics Division at the Oak Ridge National lab. \FredNote{What is the status of this group?}

Finally, we will grow our ongoing collaboration with the NSF Institute for Mathematical and Statistical Innovation (IMSI) %, hosted at University of Chicago 
(5 miles south of our campus). %\marginpar{\color{red}wait, we don't have a letter from IMSI. but it's different we are just gonna propose events there, right?} 
Our RTG trainees will participate in IMSI programs, and we will continue to co-sponsor programs of mutual interest with IMSI. 

Letters of collaboration from these partners are included as supplementary documents.







%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Proposed Project } 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 

\subsection{Research Thrusts} \label{sec:researchproblems}



The key faculty and scientists for each thrust are listed below. 
% this discussion edited to better reflect the proposal/personnel structure 
The structure of our department is similar to the way the research thrusts interact, tied together by fundamental mathematical constructs rather than broad research areas. \FredNote{I do not understand this previous sentence.} As such, there is a natural team structure for each thrust:  at least one of the PIs,  additional researchers from the department, additional researchers either from a partner outside our department, % The team for each Thrust will consist of one or more PIs, researchers with at least one Argonne scientist or faculty member from a department outside applied mathematics, {\color{magenta}this is NOT yet true for sonja's Thrust. %Argonne can collab only if application of sampling to optimization!}  
plus applied mathematics undergraduate and graduate students and postdoctoral fellows.
%{\color{cyan}New strength:}
The core of our team for the proposed project includes NTT and TT faculty across all levels in our department. The planned activities synergize across undergraduate, graduate, and postdoctoral training and collaboration and research. They are planned so as to include, invite, and impact all students, postdocs, and faculty in the department, in addition to RTG trainees.
%directly supported by this RTG. 
%In addition to those trainees (postdocs and students) supported directly by this RTG, other students, postdocs, and faculty members will be encouraged to join these teams. 




%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Thrust $1$. Variational Analysis and Inference of High Dimensional Complex Dynamics}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Chun Liu (lead), Trevor Leslie (Illinois Tech, Applied Math), Xiaofan Li (Illinois Tech, Applied Math), Yubin Lu (Illinois Tech, Applied Math), Georgia Papavasiliou (Illinois Tech, Biomedical Engineering), and Ming Zhong (Illinois Tech, Applied Math)

\subsubsection*{Background} One of the most crucial tasks in data science is the capability of dealing with high-dimensional data efficiently and effectively, such as high-resolution image generation, large language models for natural language processing \FredNote{You are talking about large language models but who has expertise?  No CS people here.}, text/image/voice mixed data analysis, medical image/data processing, etc.  Many of high dimensional data sets which we are interested in obey some form of physical/biological/chemical laws governed by systems of ODE/PDE, which makes the analysis and inference extremely difficult.  We propose several energetic variational approach based data-driven methods to infer the governing equations from observation data for analysis, prediction, and control.  The framework of energetic variational approach, originated from seminal works of Raleigh \cite{strutt1871some} and Onsager \cite{onsager1931reciprocal,onsager1931reciprocal2}, provides a paradigm to determine the dynamics of a non-equilibrium system from prescribed energy-dissipation laws.  Such paradigm offers a unique pathway into the accurate scientific discovery of models from high-dimensional dynamical data. 

\subsubsection*{Analyzing Complex Behaviors from High dimensional Data}
Complex systems of interacting agents (i.e., particles, cells, bacteria, animals, human beings, robots, planets, etc.) %\footnote{Agents here can be referred to as particles, cells, bacteria, animals, human beings, robots, planets, etc.} FOOTNOTE TAKES TOO MANY LINES OF SPACE FOR NO GOOD REASON.
can produce a wide range of interesting and intriguing patterns.  Among these complex patterns, collective behavior (such as clustering, flocking, milling, swarming, synchronization, etc.) is one of the most fascinating and difficult to analyze and model mathematically.  Furthermore, proper modeling and control of such behaviors can have significant contribution and application in understanding super conductivity, cell aggregation, pigment/fingerprint formation, social behaviors in insects (bees, ants, and locusts) and human, market behavior, opinion/consensus formation, social network analysis, etc. We employ the energetic variational approach to analyze complex collective behaviors such as line-formation (emergence of leaders in social sciences), swarming, multi-modal emergence of several behaviors, and stochastic patterns.  This analysis can be done by investigating the corresponding energies responsible for producing the patterns and derive the stability, regularity and steady states of such systems.  Proper analysis of new collective dynamics can also be used as a physical prior in the data driven modeling of collective behaviors.

\subsubsection*{Inferring Complex Behaviors from High dimensional Data}
Data driven modeling of complex behaviors is also critical in understanding collective behaviors, as it can provide validation as well as improvement to existing models by inferring physically meaningful governing equations from observation data.  Hence, another important application of our energetic variational approach is to derive accurate and physical meaningful governing equations from observation data.  Various machine learning methods have been developed in the group for data-driven modeling of collective behaviors from observation data~\cite{LZTM2019, ZMM2020, MMQZ2021, FMMZ2022,MTZM2022, ZMM2023}.  We aim to provide new learning methods for dealing with more complicated systems, complex observation schemes, and limited knowledge of labeled data.

\subsubsection*{Other Directions}
Moreover, the team has the track record and relevant expertise to lead other interdisciplinary projects, such as those related to the modeling of active soft materials in biological environments.  The interdisciplinary nature of the team will provide an opportunity to heave balanced training in difference approaches of applied math, most notably multi-scale modeling and analysis associated with the applications in biology and soft matter physics.  The energetic variational formulation can also provide guidelines to develop numerical schemes that preserves the basic physics, such as frame-indifference, positivity of physical quantities and conservatino of mass, for many complicated systems arising from multi-scale multi-physics problem in physics, biology and material science \cite{liu2020variational, liu2020lagrangian, liu2020structure}.  The same numerical framework can also be applied to develop new algorithms for machine learning \cite{wang2020particle}.  These projects will provide training in the areas of scientific computation, multi-scale simulation and machine learning.
% We focus on new mathematical modeling, nonlinear partial differential equations, scientific computing and machine learning. The framework of energetic variational approach, originated from seminal works of Raleigh \cite{strutt1871some} and Onsager \cite{onsager1931reciprocal,onsager1931reciprocal2}, provides a paradigm to determine the dynamics of a non-equilibrium system from prescribed energy-dissipation laws.  These approaches have motivated much work in the area where it is crucial to preserve the underlying thermodynamics structures, including quantification of noises and fluctuation of the systems.  The variational principle has been employed to study many complex fluids, such as liquid crystals, two-phase flows, and ionic fluids, which involve the coupling and competition of various mechanical and chemical mechanisms in different scales \cite{lin2001static, feng2005energetic, LiLiZh05, Lin2007, liu2009introduction, du2009energetic, sun2009energetic, eisenberg2010energy, Giga2017, Liu2019, knopf2020phase}.
% In an isothermal closed system, an energy-dissipation law, coming from the first and second law of thermodynamics, is often given by \dd\ddtEtotal=−△\frac{\dd}{\dd t} E^{\rm total} = - \triangle,
% where EtotalE^{\rm total} is the total energy, including both the kinetic energy K\mathcal{K} and the Helmholtz free energy F\mathcal{F}, and △≥0\triangle \geq 0 is the rate of the energy dissipation which is equal to the entropy production in the situation. The energy-dissipation law, along with the kinematics relation, %describe all the physics and the assumptions for a given non-equilibrium system. 
% capture all specific physical and biological properties for a given non-equilibrium system. Starting with an energy-dissipation law, the EnVarA framework derives the dynamics of the systems through two variational principles, the Least Action Principle (LAP) and the Maximum Dissipation Principle (MDP). The LAP, which states the equation of motion for a Hamiltonian system
% can be derived from the variation of the action functional A=∫T0[K−F]\ddt\mathcal{A} = \int_{0}^T [\mathcal{K} - \mathcal{F}] \dd t with respect to the flow maps, gives a unique procedure to derive the conservative force for the system. The LAP is indeed a manifestation of the rule δE=force⋅δx\delta E = {\rm force} \cdot \delta x. The MDP, variation of the dissipation potential D\mathcal{D}, which equals to 12△\frac{1}{2}\triangle in the linear response regime, with respect to the rate (such as velocity), gives the dissipation force for the system. In turn, the force balance condition leads to the evolution equation to the system, δDδ\xt=δAδ\x.\frac{\delta \mathcal{D}}{\delta \x_t} = \frac{\delta \mathcal{A}}{\delta \x}. Through different choices of the free energy F\mathcal{F} and the dissipation △\triangle, the EnVarA systematically deals with the coupling and competition of mechanical, chemical and thermal mechanisms in different time scales in a thermodynamically consistent framework. It provides a unified framework to study various dynamical systems demonstrated in the other projects, and moreover, gives a way to quantify energy transduction between different processes.  
% \subsubsection*{Applications in Biology and Engineering}
% The team has the track record and relevant expertise to lead interdisciplinary projects. We plan to focus on projects related to the modeling of active soft materials in biological environments. 
% The unified framework will provide the insights for the systematical
% study of coupling and competition of these multiscale and multiphysics systems.
% The interdisciplinary nature of the team will provide an opportunity to have balanced training in different approaches of applied math, most notably multiscale modeling and analysis associated with the applications in biology and soft matter physics.
% \subsubsection*{Physics Preserving Numerical Simulations}
% The energetic variational formulation also provides guidelines to develop numerical schemes that preserves the basic physics, such as frame-indifference, positivity of physical quantities  and conservation of mass, for many complicated systems arising from multiscale multiphysics problem in physics, biology and material science \cite{liu2020variational, liu2020lagrangian, liu2020structure}. The same numerical frame can also be applied develop new algorithms for machine learning \cite{wang2020particle}. Different spatial discretizations, including finite element, spectral methods, particle methods and neural network discretization, and time-stepping techniques can be incorporated in the numerical framework. These projects will provides training in the areas of scientific computation, multi-scale simulation and machine learning.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Thrust 2. Sampling Combinatorial Structures} 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 Sonja Petrovi\'c (lead),  with colleagues from Illinois Tech, Applied Math: 
 Robert Ellis,
  Hemanshu Kaul, %(Illinois Tech, Applied Math), 
 Michael Pelsmajer, %  (Illinois Tech, Applied Math), 
and  Despina Stasi; % (Illinois Tech, Applied Math). 
and collaborators from Argonne National Laboratories led by Sven Leyffer.  
 %  TRACK RECORD + CROSS-DISCIPLINARY COLLAB = PLAN ..  %% 
%\marginpar{Unfortunately, SP is the only PI on this list - the rest are senior personnel :(} 
% entire Thrust 2 should be 1.5 pages! 

The research encompassed by this Thrust  is guided by 
% challenging applications across different areas, all of which pose common challenges: (a) they involve a large number of variables or combinatorial constraints; (b) they are encoded as high-dimensional complex discrete structures; (c) they cannot be solved with current state-of-the-art methods.  The 
structured  computational problems that are critical to discrete optimization, computational algebra, and statistics, and their solution is amenable to randomization techniques. 
%In a tightly integrated research program, we propose to (i) develop new structured sampling techniques that exploit combinatorial structure; (ii) investigate new randomized algorithms based on efficient structured sampling for  inference, model fit testing, and discrete optimization problems; (iii) 
We will leverage our team's track record, ongoing collaboration, and  combined expertise in discrete and computational mathematics, statistics, and randomized algorithms to develop  blended hybrid approaches to the following three  problems. 
%In particular, we will focus on the following three areas. 


\subsubsection*{FitNet: information-preserving data reduction,  networks, sampling, and estimation} 
 Statistical network models for discrete relational data  have grown in popularity in recent decades, as illustrated in \cite{goldenberg2010survey} \marginpar{\color{red}why is my citation not showing up? it IS in the bib file.} and \cite{KolaczykBook2017}. They are designed to model various types of interactions between a set of entities--nodes in the network--such as  individuals \cite{JiJinAuthorsAOAS}, organizations \cite{petrovic2019}, proteins or neurons \cite{GPS21+}, or brain regions \cite{karwa2016exact}. Links represent observed relationships between the nodes, such as personal or organizational affinities, social or financial or economic relationships, chemical or physical links, and can be directed or undirected, or be counted with  weight. Social networks may include nodes’ gender, education, income, etc.; these   are often of key importance for causal inference. 

The question of model goodness of fit is central in statistics and unsolved in data science; it is critical for   estimating network features, appropriately simulating data,  and correctly interpreting the results.  %``Doing data science" does not make sense without a proper way to test model fit, as using wrong models will simply lead to wrong conclusions. This step is separate from what is often referred to `validation' in practice. 
It is also considered to be very challenging in the network science community due to both the size of the networks and their structure. Our  hybrid algebraic-combinatorial approach  \cite{GPS16} has proved to be efficient, solving a decades-old problem in statistics by developing a dynamic construction of  an irreducible Markov chain for model testing. This method can be applied to networks with a thousand nodes, while state-of-the-art software \cite{4ti2} that directly computes Markov bases can handle networks with at most 10 nodes because it is, by design, data agnostic. 
%PI Petrovi\'c has extended this dynamic sampling methodology to other problems in statistics [65, 79].     

During the proposed project, we  are specifically interested in a new sampling framework that uses a parameter hypergraph, which can be abstracted by building on PI Petrovic’s work, an important  part of which was coauthored with the team on this RTG \cite{shellERGM,betaHypergraphs,tergms,bouquetAlgebra,KarwaPetrovic:AOASauthors,Psurvey,PTVhypergraphs,RPF:11}.  A Masters thesis she supervised in 2016 \cite{DillonMS16} derives a rapid mixing result for the model fit Markov chain. This proposal will generalize this result to other models. %% THIS WAS A **DOMESTIC STUDENT**
%{\color{magenta} I fail to provide any mathematical details. this is simply very broad.} 

 \subsubsection*{Dynamic lattice sampling and discrete optimization.} 
 Petrovi\'c, Leyffer, and current PhD students and postdoc are developing sampling techniques that exploit combinatorial structure in order to develop randomized algorithms for massive discrete optimization problems that are intractable with other solvers. In joint work with PhD student Bakenhus, PI Petrovi\'c tackles the problem of sampling the set of nonnegative integer lattice points in a polytope. This set, also known as the fiber of a linear map, is the support set of a (mixed) integer program. We propose a biased sampling algorithm with a Bayesian framework, such that  the bias embedded in our algorithm updates sampler parameters to improve fiber discovery rate at each step chosen from previously discovered elements. As of this writing, we are able to efficiently (on a laptop) sample from fibers that are entirely out of reach for the state-of-the-art Markov bases samplers. 

In ongoing work, the Argonne team led by SP Leyffer have considered a mixed-integer robust optimization model expressible as a problem supported on a fiber  for identifying potential weaknesses in an interconnected infrastructure setting.  The state-of-the-art discrete optimization solvers face fundamental challenges when tackling problems of this form. In our joint work, we will develop \emph{randomized} discrete optimization methods to tackle a new class of challenging societal problems. Our prior joint work, e.g., \cite{PShypergraphs}  highlights how the clever use of sampling algorithms to dynamically create polynomial bases for sampling lattice points in polytopes fibers can turn an impossible- to-scale algorithm into a dynamic, data-driven, scalable one of practical use.


\subsubsection*{Dynamic networks}\marginpar{\color{magenta}TAG: Hemanshu, Michael - Sonja sent an email. please add citations to this part if you feel like.}
Finally, we will investigate {\it dynamic networks}: networks that evolve over time, or networks where each interaction happens at a specific time.  
SPs Kaul, Pelsmajer, and Stasi will co-lead an  investigation into appropriate adaptive time scales  for summarizing dynamic networks and develop corresponding measures of quality. 
Aggregation of longitudinal interaction data capturing multiscale phenomena will be combined with methods from statistics and network theory to be able to handle both temporal errors (sampling/data collection errors) and topological errors (missing or spurious interactions in the network).

The longitudinal data of interactions in such networks has to be carefully analyzed to discover persistent patterns, such as hidden communities in a dynamic network of social interactions. Traditionally in graph theory, such data is aggregated into a single network, losing the temporal aspect of the data and also making its size difficult to analyze simultaneously. Our recent joint work \cite{tergms} makes a first connection between this idea and statistics. Another way is to take ``snapshots" of the dynamic network data at appropriate times where the timing of such snapshots can make a dramatic difference in capturing or missing the essential aspects of the network behavior. 

In order to summarize the dynamic network data, one must determine an appropriate time-scale, or adaptive time-scale, or multiple time-scales to capture multi-scale phenomena. The aggregation of this data based on such time-scale(s) must be combined with methods from statistics and network theory to be able to handle both temporal errors (sampling/ data collection errors) and topological errors (missing or spurious interactions in the network).  We will develop network algorithms and statistical models to handle this data time-scale aggregation problem in order to capture more types of network behavior and to do it better.  We will  consider different measures of quality and evaluate the success of our methods and models on  real and synthetic dynamic network data.






%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Thrust 3. Dynamics  Under Uncertainty } 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\marginpar{this one isn't first, right? this is Thrust 3.} 
Xiaofan Li (lead), Fred Hickernell, Michael Pelsmajer, Ming Zhong (Illinois Tech, Applied Math), Mustafa Bilgic (Illinois Tech, Computer Science),  and Prasanna Balaprakash (Argonne National Lab). 

\subsubsection*{Background} The interactions of uncertainty and nonlinearity lead to intriguing phenomena, such as  transitions between  different dynamical regimes. We shall develop modeling, analytical, computational and data science  methods to quantify stochastic dynamics, with applications to chemical \cite{agaoglou_chemical_2019}, biophysical \cite{Ruoff2018BiologicalCR}, power grid \cite{MEDJROUBI201714} financial risk \cite{DixHaBil20a,Gla03}, and climate systems \cite{Alexandrov2020NonlinearCD, Franzke2017NonlinearAS, Wan2020ADF}. 
 
Complex systems are usually under the influence of randomness or uncertainty \cite{Moss1, Horst, Gar, VanKampen3}. Uncertainties may also be caused by our lack of knowledge of some physical processes that are not well represented in the mathematical models (epistemic) and due to the inherent randomness of a specific event  \cite{Palmer1, Kantz, Wilks, Williams}.
Although these random mechanisms appear to be very small or very fast, their long time impact on the system evolution may be delicate or even profound \cite{Arnold, DuanBook2015}. 

%These delicate impacts on the overall evolution of dynamical systems have been observed in, for example, stochastic bifurcation \cite{Crauel, CarLanRob01, Horst}, stochastic resonance \cite{imkeller2002model},  and  noise-induced pattern formation \cite{Gar, blomker2003pattern}. Hence, taking stochastic effects   into account is of central importance for mathematical modeling of complex systems under uncertainty.   
%It is therefore crucial to investigate dynamics under uncertainty, in the context of models arising from applications in, for example, geophysical and biophysical systems. 

In the absence of fluctuations, a physical system might be modeled by an  ordinary differential equation (ODE): $ dx/dt=f(x)$.  However, random fluctuations are often significant in practice. While Gaussian disturbances, say driven by a  Brownian motion, $B_t$, may suffice in some situations, they may not capture the true dynamics.  A better choice is often \emph{non-Gaussian} fluctuations, e.g., represented by a L\'evy motion $L_t$ \cite{Woy,Dit, Shlesinger,taqqu,dybiec2009levy}. For instance, it has been argued that diffusion by geophysical turbulence \cite{Shlesinger}
corresponds  to a series of  ``pauses", when the
particle is trapped by a coherent structure, and  
``jumps" (a transition event) to another regime, when the particle moves in the jet
flow. Paleoclimatic data \cite{Dit, ZhengYY2020}   indicate   sharp transitions in temperature as in L\'evy fluctuations. Gene transcriptions have also been studied as noise-induced transitions \cite{Raser2005,Suel06,Lin2016GeneED, ChenWuDuan}. 
%A further example is a thermally activated motion of a test particle along a polymer,   shown in \cite{sokolov1997paradoxal, brockmann2002levy} to be subject to 12\frac{1}{2}-stable L\'evy motion due to polymer's self-intersections. 
In these cases, we should use a
 stochastic   differential equation (SDE)  model \cite{Arnold,   Oksendal, DuanBook2015, Applebaum}
 driven by a L\'evy motion as well as Brownian motion:
 $
 dX_t= f(X_t) dt + g(X_t) dB_t + \sigma(X_{t-}) dL_t.
 $
 A physically significant L\'evy motion is the  $\alpha$-stable L\'evy motion $L_t^\alpha$ (for $0<\alpha<2$).  
 
% We often think of this as the governing equation for random `particles'. 
 
Stochastic dynamical systems are also extremely relevant to data science applications such as machine learning in the presence of  noise. 
The study of complex stochastic dynamical systems with a focus on their interface with state-of-the-art deep learning architectures will prepare applied mathematics undergraduate and graduate students for a wealth of opportunities in data-driven modeling and simulation of problems with high societal impact, such as climate change, gene regulation or mutation, and shocks to financial markets.

%Recently, ODEs and SDEs parameterized by neural networks have been used for learning regression and classification tasks for noisy irregular datasets \cite{chen2018neural,rubanova2019latent,jia2019neural,tzen2019neural,look2020deterministic,liu2019neural}. Specifically, SDEs, with non-Gaussian as well as Gaussian noise, have been used to represent the `flow' of information from the input of a deep learning framework to the output. The assumption of noise-based forcing of dynamics imparts significant robustness to the quality of learning and the ability to generalize to unseen data or encrypted data (especially relevant to the post-quantum era, linking to Thrust 3). 

\subsubsection*{Quantifying transition phenomena}
To understand transition in a stochastic system, we study the `ensemble' (or `continua') quantities: mean exit time (exiting from a dynamical regime), escape probability (from one dynamical regime to another), and  the transition  probability density. Ensemble quantities help us better understand  climate shifts \cite{ZhengYY2020} and  gene transcription \cite{ChenWuDuan}. 
The ensemble approach leads to deterministic  nonlocal (integro-differential) equations for mean exit time and for escape probability, and a nonlocal Fokker-Planck equation for the probability density \cite{DuanBook2015},
in which the usual Laplace operator is replaced
by a nonlocal Laplace operator, which is related to the generator of L\'evy motion $L_t^\alpha$:  $-(-\Delta)^{\frac{\alpha}2  }$, for $\alpha \in (0, 2)$.
 
% to simulate these high dimensional nonlocal equations based on a stationary or temporal normalizing flow model, as a  generative machine learning tool.  This new method 
 
As the usual numerical methods  are not feasible, we will develop a method with techniques to handle the nonlocal terms implicitly since arbitrary probability densities can be learned from samples alone. Subsequently, `ensemble' quantities, including the mean trajectories (when they exist), are high dimensional integrals with respect to the probability density \cite{DuanBook2015}.  Thus we will validate our fit distributions with Monte Carlo simulations, in particular highly stratified or quasi-Monte Carlo (QMC) variants, where our team has made recent theoretical advances \cite{Hic17a, HicEtal17a} and efficient implementations \cite{QMCPy2020a,ChoEtal22a}.

Furthermore, we will investigate the most probable transition pathway between dynamically different regimes (e.g., metastable states) during finite transition time. This pathway is the minimizer of the associated Onsager-Machlup action functional, $\mathcal{A}(z, \dot z) = \int_0^T \mathcal{L}(z(t), \dot z(t)) dt,$ where the Lagrangian  $\mathcal{L}$ depends on the vector field $f$, noise intensities $g, \sigma$, and  jump measure $\nu$ for L\'evy motion; for background, see   our recent works \cite{ChaoDuanOM,HuangYF}. The   transition time $T$ may also be estimated \cite{HuangYF2020}. We will explore how the energetic variational approaches in Thrust 2 may benefit us in the minimization of the Onsager-Machlup action functional. 

QMC simulations are often used to quantify financial risk as well as in the problems mentioned above.  We will construct reliable data-based error estimates for QMC, which can serve as stopping criteria for QMC algorithms.  Although PI Hickernell and his collaborators have made some progress \cite{HicJim16a,JimHic16a,HicEtal17a,RatHic19a,JagHic22a}, error estimates for very high-dimensional problems treated by multi-level QMC is an important open problem.

\subsubsection*{Extracting transition phenomena from noisy data}
Some   stochastic systems are known or partially known (some mechanisms are missing) via noisy observations or simulated data sets. We will  develop techniques to extract stochastic governing laws \cite{Wei2022AnOC, YangLi2020a, Li2022ExtractingGL, Lu2022LearningTT} and then examine transition phenomena by computing mean exit time, escape probability and transition probability density as above. 
There are several potential interfaces with state-of-the-art generative machine learning, which has been used to design novel methods to learn probability densities from data via neural stochastic differential equations. Furthermore, our colleagues'  work on 
designing and optimizing sensor networks, and graph theory \cite{karwa2016statistical,Calines2008MonitoringSF} will be explored to guide smart data collection for interface with such novel algorithms. 

%A class of high dimensional dynamic systems may be interpreted as networks, such as   gene regulation networks \cite{Raser2005} which are modeled by systems of stochastic differential equations \cite{Suel06}.  In a gene regulation network, a transition from low to high concentration of a promoter protein may indicate transcription \cite{Stefan,ZLDK}. Studying transitions in large networks is also of vital importance for resiliency analyses and control of power grid infrastructure and for supply chain modeling in operations research which may be modeled as stochastic processes on graphs \cite{shin2020graph,anghel2007stochastic,nardelli2014models}. For large networks with multiple sources, sinks and connectivity measures of several variables, extremely high dimensional systems are commonly obtained. Furthermore, optimization of these graphs for desired quantities of interest require the solution of a complex mixed-integer nonlinear programming problem with constraints that may not be nominal input-output mappings but aforementioned complex dynamical systems \cite{shin2020decentralized,sampat2017optimization,kim2019graph,shin2021exponential}.


\iffalse
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{\color{magenta}DELETE THIS ONE Thrust 3. Mathematical Tools for Open Quantum Systems}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Jinqiao Duan (co-leader), Fred Hickernell, Xiaofan Li,  Carlo Segre (co-leader, Illinois Tech Physics), Romit Maulik (Argonne National Lab \& Illinois Tech) and Martin Suchara (Argonne National Lab).

\subsubsection*{Background} The development of quantum science and its exploitation in technology is a topic of great interest and activity in many countries. The United States \cite{raymer2019us}, Europe \cite{riedel2019europe}, and China \cite{kania2018quantum} have invested heavily in long term ``quantum initiatives’’.  It has been stated that we are in the midst of the ``second quantum revolution’’ \cite{kania2018quantum}. Indeed,  most people tend not to realize just how much quantum mechanics influences their day-to-day life. For example, consider that standard household light bulb. Advances in light emitting diode (LED) technology have led to ``white’’ light bulbs that can now be purchased anywhere light bulbs are purchased. These LED light bulbs are more energy efficient, last longer, and are more environmentally friendly than light bulbs of the past. This is all thanks to quantum mechanics, which is fundamental to LED physics and technology. The impact of this rather mundane example is significant. But the impact of more complex physical and technological quantum advances, such as quantum computing, quantum cryptography, quantum sensors \cite{ng2020guest}, and even quantum artificial intelligence \cite{taylor2020quantum} will change the world and our way of life.

Therefore it is essential in education nowadays that students develop a level of ``quantum literacy’’ \cite{foti2021quantum} that will enable them to interpret and evaluate these new phenomena and technologies in a way that will enable them to assess their impact, not only on their own  lives, but on society and the world. Developing this quantum literacy in the context of mathematical education,    at both the undergraduate and graduate   levels,  is a goal of this RTG project.


Quantum mechanics is not typically emphasized in mathematics programs.  Indeed, many of the unusual properties of quantum mechanics, such as tunneling, entanglement, superposition of multiple states, interference of states, uncertainty, and quantum measurement can be understood from the fundamental physical principle of ``wave-particle’’ duality and de Broglie’s relation between the momentum and the wavelength of a particle.  From this one is led to the Schr\"{o}dinger equation description of particles in terms of waves. In fact, the physical setting of quantum mechanics is described by a few (mathematical) axioms that describe the state space, observable quantities, measurement, and time evolution.  Once this physical setting is translated into mathematics, the implications and consequences of the mathematics is a natural setting for the participation of mathematicians. However, the required mathematical skills and knowledge are not typically emphasized or even covered in relevant courses. This RTG will remedy the situation, both through training undergraduate, graduate, and postdoctoral students, and through the development of a course appropriate for mathematicians, but which bridges this gap. This course will lead naturally into a variety of research areas that fit naturally into the context of the other research topics, that we will describe. Indeed, we will explore how   Thrust 3 helps us in machine learning of encrypted data, especially for post-quantum era, linking with Thrust 1.  

\subsubsection*{Exploring the Interplay Between Quantum Mechanical Uncertainty and Noise.}  Environmental noise   arises naturally in the mathematical formulation of open quantum systems \cite{Lidar2019LectureNO, Breuer2002TheTO,Macieszczak2021TheoryOC, Griffiths2018IntroductionTQ}. This Thrust will explore the interplay between quantum mechanical uncertainty and the influence of external noise \cite{Lindgren2019QuantumMC, Nagasawa2000StochasticPI} on the quantum mechanical system, including metastable transitions \cite{Macieszczak2021TheoryOC, Rose2019MetastabilityRA} and the connection with our earlier works on the most probable transition pathways in stochastic systems with non-Gaussian L\'evy noise \cite{ChaoDuanOM}. Moreover, we will examine how certain quantum dynamical behaviors  may be understood through stochastic minimization \cite{Lindgren2019QuantumMC} and affected by quantum noise \cite{Nurdin2019QuantumSP}. This may   benefit from our recent work on effective quantum wave factorization \cite{ZHANG2020132573}. This also relates to   Thrust 1 - Dynamics under Uncertainty.


\subsubsection*{Applying Methods of Lagrangian Fluid Transport to the Bohmian Trajectory Formulation of Quantum Mechanics.}   A `trajectory'  formulation of quantum mechanics  equivalent to the Schr\"{o}dinger equation can be given \cite{Bohmian, Holland1993TheQT} that emphasizes the flow of probability, or ``probability current’’ in space. 
In this formulation, the state of a system of NN quantum particles is described by its wave function ψ=ψ(q1,…,qN)=ψ(q)\psi=\psi(q_1,…,q_N)=\psi(q) via the Schr\"{o}dinger equation  on the space of possible configurations qq of the system, together with its actual configuration QQ defined by the actual positions Q1,...,QNQ_1,...,Q_N of its particles. The trajectory of these quantum particles obey the dynamical system
$
\frac{dQ_k}{dt}= \frac{\hbar}{m_k} \text{Im} \frac{\bar\psi  \partial_k \psi}{|\psi|^2}(Q_1,…,Q_N),\;\; k=1, ..., N,  
$
where ℏ\hbar is the reduced Planck constant and mkm_k is the mass of kk-th particle.
This is the pilot-wave, or hydrodynamical, point of view. We will explore this point of view from the perspective of the dynamical systems approach to Lagrangian transport in fluid dynamics, including an examination of phase portraits of quantum dynamical trajectories \cite{Berndl1995OnTG}, transition phenomena \cite{waalkens2007wigner,Micha2006QuantumDW, Dittrich2016}, and effective approximation inspired by our recent work \cite{Li2021BohmianTO}. 
\fi


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Regular Research  Activities}\label{section:regular research activities}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% The groups of faculty, postdocs, and students corresponding to the three research Thrusts will meet regularly during the academic year and the summer. In addition to the focused research activities on problem-solving, the interactions will include department-wide seminars. The PIs already run a
Each thrust from Section 2.1 represents new interdisciplinary connections between our faculty, which will be used to strengthen department-wide activities. The PIs already run  research seminars: 
Stochastic and Multiscale Modeling and Computation seminar for Thrust 11, Nonlinear Algebra and Statistics (NLAStats) and Discrete Math seminars for Thrust 22, and a Computational Mathematics and Statistics seminar for Thrust 33.
In order to facilitate cross-thematic collaboration, the seminar times will be scheduled to not overlap. 


%THIS SUBSECTION DESCRIBES *REGULAR* RTG ACTIVITIES, BUT THEN WE ALSO HAVE SUMMER AND OTHER CO-CURRICULAR TRAINING ALL OF WHICH ARE ALSO "OUR ACTIVITIES" SO THIS IS JUST A STUPIDLY ORGANIZED SECTION! THIS REALLY SHOULD NOT BE activities but day-to-day nuts-and-bolts!!! "research activities"!!!

%{\color{magenta}Sonja will add: student training will incorporate  statistics and data literacy.} 

In terms of other day-to-day activities by the research group, 
%\subsubsection*{Nuts and bolts} 
%\marginpar{\color{magenta}this paragraph shall become <10<10 lines.}
all trainees (students and postdocs) supported by this RTG will be assigned to one or two projects and join the corresponding seminar. 
Trainees may enter or exit these research groups at different times of the year. Each seminar will include background lectures. To ensure  continuity and  help connect outgoing, current, and incoming trainees, background lectures will be  %, e.g., some will participate during a summer only, while others will participate over the course of a couple of academic years. Thus, certain background lectures will need to be repeated periodically.  
initially given by the PIs and our faculty and scientific collaborators, and the lectures will be recorded so that students and postdocs can refer to them.  However, part of our training will be to mentor our postdocs and experienced students to give these background lectures as well.  
Meetings will be held in hybrid mode as necessary. %much as possible to allow , i.e., in person and via video conference, to allow the greatest participation.  {\color{magenta} youtube broadcast?!\sout{Workshops, tutorials, and public lectures will be broadcast via our Illinois Tech YouTube channel to allow greater access, both synchronously and asynchronously.}}
Trainees entering the research groups will be provided with a faculty mentor and a postdoc or experienced student mentor.  New trainees will be given an article to read related to the research problem and report on it to the seminar, describing what are the significant claims, pointing out drawbacks, and suggesting directions for extension. The articles may be in mathematical sciences or in an application area.
As trainees become engaged in research, they will report on their progress at the regular seminar meetings, while substantial new results and new papers will be presented at regional and national conferences and in refereed publications.



Members of research groups will actively apply for external funding to further support their research activities.  PhD students and postdoctoral fellows will be asked to draft parts of the research proposals and also to read and comment on draft proposals as part of their training.  BS and PhD students will be encouraged to apply for graduate and postdoctoral fellowships from the NSF and other external funding agencies.  As the teams collaborate, they will be in a better position to compete for  non-traditional and interdisciplinary funding opportunities.  Their successes in these awards will support these research activities beyond the life of this grant.

\subsection*{Applied Math, Data Science, and Social Responsibility.} 
% The point here is that SoReMo is recurring, and we will make the math students connect with it and be very RESEARCH ORIENTED. So yes it is broader impact but it should be research, first and foremost, like Chun had said before. So this description better belongs here. 
PI  Petrovi\'c  leads a new, university-wide initiative called Socially Responsible Modeling, Computation, and Design (SoReMo).  It advocates for ethical, equitable approaches in computation, modeling, and design, contributing to the common good of Chicago and beyond through research and education initiatives at Illinois Tech. As there is \emph{no} sustained funding for this program, each department or college supports is when they are able. Hence this RTG will allow us to  train applied mathematics students with the projects relevant to SoReMo's mission.  Fellows are all of the following: idea creators, proposers, researchers, and deliverers. The faculty will guide the students or student teams throughout the semester. Undergraduate students will participate as standard SoReMo Fellows and will be trained on writing project proposals (see Section 2.4 for details). PhD students will be involved as special Data Science Fellows,  a new role we will create to connect 
\marginpar{\color{magenta}Rob, FYI: I have renamed them from "bridge fellows" to "Data Science Fellows".} 
%ROB ends this paragraph here. Please add a blurb :) } \marginpar{\color{magenta}TAG: Rob! I-course. Thanks Rob-  this is EXCELLENT! }
%{\color{magenta}FROM ROB:} DS 480 (formerly Math 497) 
SoReMo activities with DS 480 Data Science Projects, which is SP Ellis’s team projects capstone course for diverse majors with sufficient preparation in computer science, math, or statistics.  The fellow will help develop early-stage SoReMo projects into full DS 480 team projects, propose follow-on projects from DS 480 outcomes for later SoReMo proposals, and mentor SoReMo Fellow applicants in their proposal preparation.  The DS 480 projects are supported by an outside expert, explicitly include ethics and broader impacts in deliverable rubrics, and usually have immediate civic relevance.  Recent topics have included evaluating and visualizing lead levels in Chicago drinking water (with Elin Betanzo of Safe Water Engineering) and congressional redistricting (with Emily Webber of Amazon Web Services).  The Bridge Fellow will help to: develop the course topic, prepare the data set(s), engage an outside expert, deliver just-in-time tutorials, coach teams, provide individual logbook feedback, and capture final work products for external dissemination.  This “super-TA” role was successfully piloted in Spring 2019, and will prepare the Data Science Fellow to run their own similar course, or lead a data science project, later in their academic or industry career.  


% WE DO NOT NEED THIS PART - IT DOES NOT ADD TO THE PROPOSAL. IT LITERALLY JUST SAYS WHY SOREMO IS DOABLE IN ITS BROAD SCOPE! :) I KEEP IT HERE AS COMMENT.
% The impact of this project on students is not in what we ask, but in \emph{how} we ask it. Our operating philosophy is horizontal, aligning with the vision of the College of Computing in particular. We seek out, identify, and help students develop social responsibility in the context of various majors across campus. They spend time working on open-ended research projects which require them to connect concepts from within and outside their coursework, connect to the community, seek out best practices and propose potential solutions to the problem at hand. 



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Summer Program}    \label{sec:summer}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Our planned summer research and training activities will be threefold: an undergraduate summer research program, summer school workshop for graduate students and postdocs, and research workshops in collaboration with institutes. To ensure program coherence and continuity, a faculty member will oversee each of the research activities each summer, jointly with a postdoctoral fellow and one or two PhD students who will help mentor the undergraduate students.  There will be a synergy between our summer program and the related research groups, which will continue to meet during the summer. 

%\subsection*{Undergraduate summer program.}
%\sout{Moreover, we will be running a three-year NSF-funded ten-week undergraduate summer research program. {\color{red}unclear. "we will be" - as part of RTG, right?organize.} The program will be targeted at rising third and fourth-year BS students.  The topics chosen each year will come from one or two of the research projects in Section ??????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????\ref{sec:researchproblems}.  We will support eight students each year, with around two-thirds being from Illinois Tech and the remaining being from other institutes.  When recruiting students from other institutions, preference will be given to women and underrepresented minorities.  RTG funds will pay a stipend and room and board to participate students.  This will allow students to engage with one another and their mentors outside of research time.  The ten-week program will begin with tutorials given by postdoctoral fellows and experienced students introducing a few possible research topics.  Participants will do some background reading and choose their topic, either solo, or in a small group, within the first week or two.  Regular meetings will allow participants to present their progress, get feedback from the larger group, and hone their message for an end of program presentation. The program will be hosted at both Illinois Tech and Argonne.  In case of a pandemic event, the summer program will be held virtually.}

 \YuhanNote{How many undergraduates will be recruited? Are they all from Illinois Tech? Will they split into the three research groups or just one or two of these?}{\color{blue} MZ: we plan to recuirt $10$ more for the UG, in addition to the SURE program, which will be led by the postdocs and graduate students from the RTG, but Yuhan will be in charge (overall) of them I think.  And they are only from IIT, and will be trained as potential student researchers later.}
We will recruit XX rising third and fourth-year undergraduates ---\emph{especially those from underrepresented groups}.
All undergraduate participants will be encouraged to write blogs \cite{Hig21a} during the program. The blogs will help them learn to articulate their research problem and their findings to a more general audience. The postdoctoral fellow and PhD
students will meet with students weekly. During the regular meetings, the postdoc and PhD students will communicate with undergraduate students and check on student progress. They will also hold regular office hours throughout the week to be available when these undergraduate students have questions. In addition, PhD students will hold several short classes or tutorials sessions to prepare the undergraduates for the projects.

Illinois Tech also hosts an REU site, Summer Undergraduate Research Experience (SURE)  program, sponsored by NSF-DMS.  Our summer program will collaborate with SURE in the recruitment process to identify students interested in the research topics that we are offering. In addition, we will coordinate with SURE to have bi-weekly presentations, final presentation, poster day, and field trips to SAS Insitute Inc.\ and Argonne National Lab. 

% A summer research program [SURE] (summer REU) so that graduate students, postdocs and PIs can participate in training undergraduate students (from outside of Illinois Tech) and prepare them for future research, collaboration with also IDEAL {\color{magenta}will return to edit this (Yuhan).}

A subset of  PIs and SPs will apply for additional funding for participant support to organize 1 or 2 summer schools and workshops for graduate students and postdocs, focusing on the intersecting topics between the three research Thrusts. One possible theme will be discrete methods, machine learning, and optimization, centered around Dr Leyffer's proposed mini-course (see page \pageref{mini courses}). 
\marginpar{\color{magenta}Ming et al., do you want to add a sentence on a topic?} 
The PIs have prior experience in organizing, running, and securing funding for such graduate training programs, including  PI Petrovi\'c's AMS-MRC (Mathematical Research Community) in 2016 and a stand-alone NSF-funded summer school in 2019 co-organized with SP Stasi. %{\color{red}Everyone: feel free to add others' prior experience as evidence that we can indeed do this, if you would like.} 
%{\color{magenta}Sonja: one of Sven's minicourses can be part of a summer school at IIT!} 
%so that the graduate students and postdocs can be exposure to research areas outside of IIT and participate in networking events during the time.
%{\bf WE WILL NOT PUT PARTICIPANT SUPPORT HERE BUT WE WILL APPLLY FOR CONFERENCE GRANT TO DO THIS. SAY THAT RTG TEAM WILL ORGANIZE 1 OR 2 SUMMER SCHOOLS. }

Finally, the PIs will continue their activities in organizing international research workshops; specifically, two- and five-day  research workshops, some at Illinois Tech, and others  as proposed events hosted at IMSI or Banff Internatinoal Research Station. The workshops' goal will be to both showcase the work of our trainees and bring in international experts to present and collaborate. We will pay special attention to diversity and inclusion when selecting participants of these events. 
%Each year, each research group will organize a two-day workshop, either at Illinois Tech, in collaboration with IMSI or at an international research center such as the one in Banff.  The workshops will showcase our own results to others outside Illinois Tech and also let our trainees hear from international experts. The PIs have a track record of organizing such events at both IMSI and Banff.  {\color{magenta} we need to tune this a bit. getting into Banff is highly nontrivial and it's very unlikely we can run a workshop each year. Also they are not 2 days, they are typically 5 days. IMSI - probably; but I don't know that they ever host a two-day event. they certainly do not advertise that they do. So let's think about what is REALISTIC and just write that.}
%And note that we have a track record of doing so. Sonja is organizing a 15-week program at IMSI *and* a Banff workshop in 2023, for example.}

%\sout{Summer is a great time for intense research when faculty and students alike take a break from formal coursework. }  Postdoctoral fellows will help RTG faculty to lead our  summer programs. {\color{magenta}HOW??}\marginpar{\color{magenta}TAG: Yuhan. }  

 % , the Department of Applied Mathematics and Department of Computer Science
 
%\subsection*{Graduate summer program} \FredNote{I do not understand this program and if or how it syncs with the undergraduate one.  Will our PhD students be pulled in two directions?} We will organize two-week summer programs for  graduate students, featuring (i) presentations by RTG faculty and collaborators about background and advances  in topics related to our research  projects; (ii) presentations by postdoctoral fellows and graduate students either about their research progress or describing specific articles related to the ongoing research projects. These will provide the general background information in the form accessible to undergraduate students.  A goal of   this summer program is to excite new students and inspire existing trainees in this RTG. 
% *OH, WHATEVER*
%We plan to have this summer program to be held at Illinois Tech in Years 1, 3 and 5, and at Argonne National Lab in Years 2 and 4. 


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Curricular and Co-Curricular Training}\label{section: co-curricular}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%{\bf REPHRASE THIS INTRODUCTORY PHRASE!} 
%To support their research activities, student trainees will take courses and participate in co-curricular events common to all research groups.  PhD students and post-docs will be involved in teaching and mentoring. {\color{magenta}of whom??} 

While our research seminars  typically feature early-career and student speakers, this grant will be used to institutionalize student and postdoc training  efforts and create an inclusive training environment with long-lasting impact. %(post-RTG).
In particular,  PIs and SPs will  design and implement a  series of new events, building on  ongoing efforts, which are \emph{centered around student and postdoc training}: 
%{\color{cyan}TO ADD-- IMPORTANT NOTES}: student training: ``elements of a research seminar". A series of events that will help bring our department together. These will be *new* events but will build on our past and ongoing efforts which were faculty initiated. 

\subsection*{Elements of a research seminar.} 
\marginpar{\color{magenta}TAG: Hemanshu.}
Approximately once per month, SP Kaul, jointly with another  faculty member (role rotating annually), will organize a graduate and advanced undergraduate student seminar event. This will be slightly different than a regular research seminar; students will present their work, or read and present background papers and open problems, or...{\color{magenta} Hemanshu, is this OK? please edit. reminds me of your discrete student talks... need a short paragraph. You have  lots of evidence of success with student speakers so far!} 
%2) A graduate student seminar 
%- once a month or so, keep it reasonable. 
%Includes student training: `elements of a research seminar'. 

\subsection*{Research project development workshops.} 
\marginpar{\color{magenta}TAG: Despina. Is this OK? Please edit.} 
In conjunction with the SoReMo program described above, SP Stasi will lead the development of a workshop, to take place once per semester, in which students will be \emph{trained on developing and writing short research proposals}. This will be open to all students at Illinois Tech, but aimed at undergraduates and graduate students in Applied Mathematics  and related majors. Students will be able to test their proposals by submitting their projects (team or individual) as SoReMo Fellow applications ..... The materials developed for this will be something the department can use post-RTG. {\color{magenta}and this is why I am asking FUNDS for this, so Despina can lead and create files and templates and documents we can keep using later. } We will store the materials on GitHub...e.g. https://github.com/SoReMo-org using info such as https://sondzus.github.io/SoReMo/  . 
%3) undergraduate research and project proposals - once a semester or so. 

%\subsection{Communication Skills and Career Coaching -- PUT THIS WITHIN CO-CURRICULAR}
%\subsection*{Colloquium} The colloquium mentioned above, although aimed in part at BS students, will also have a number of activities aimed at PhD students and postdoctoral fellows as well.  Some of these are mentioned below.
\subsection*{Communication Training}
Communication is crucial to a scientist, and the crafting of the message depends on the audience:  research group, journal referees, proposal evaluation panel, students, and/or the general public. We will focus on three ways of developing trainees' communication skills: 

1. General public audience.  C2ST (see the introduction) will run a yearly workshop consisting of three modules of three hours each duration on communication to the public. The topic will be: speaking to the general public, social media engagement, and writing for the general public.  %This will be a continuation of the undergraduate RTG event mentioned above but include PhD students and postdoctoral fellows as well.

2. Broad scientific audience. %{\bf Communication training.}
This part will be developed jointly with the SoReMo program. Fellows engage in project shareouts twice a month in a public Forum. They update their peers, faculty, and visiting students  about their progress, questions, issues, revelations, and stumbling blocks. We  train students on how to effectively design and deliver pitch-style project summaries.

3. Expert audience. %conferences - expert audience. 
The RTG will provide travel support to trainees to  present   and to network in local and regional activities, e.g.  the Chicago Area Undergraduate Research Symposium (CAURS), the Chicago Area SIAM Student Conference, AMS Section meetings, Midwest Dynamical Systems Conferences, Midwest Numerical Analysis Day,  Midwest Probability Colloquium, and workshops sponsored by Argonne and IMSI. Trainees presenting will practice their talk  with PIs and peers. \label{C2ST}
%{\color{red}MOVE THE FOLLOWING SENTENCE TO THE BUDGET JUSTIFICATION DOCUMENT AND DELETE FROM HERE: We estimate for RTG students to receive support to one conference per year in the final 2 years of their studies and one conference per year to postdoctoral fellows. } 


\subsection*{Publishing and writing.} Senior graduate students and postdocs will gain experience with the full publication process using our in-house formal research journal hosted by the Galvin Library. Trainees will be given opportunity to act as guest editors, referees, and production editor process participants for the SoReMo journal publication, handling the math and data science related publications that come through. The journal handles about a dozen submissions a year, so this is a reasonable workload for training purposes. PI Petrovi\'c and SP Popovi\'c will oversee this  part. 
%-- this is also communication (written) but it is more as it gives trainees FULL PUBLICATION PROCESS experience! {\color{magenta}SONJA --- rewrite this to incorporate students as guest editors, referees, production process participants, as well as authors. ---} Each fellow or team submits a final technical report, to be readable by a general audience, to our in-house journal publication—the SoReMo journal–hosted by our own Galvin Library. This is a formal publication with DOIs for each article. The process mimics a ‘real’ academic journal and the reports undergo a round of anonymous reviews and revisions before formally accepted and formatted for publication. This gift will help streamline this publication process by offloading some of the editing work to student trainees who will gain experience and training in editing and publication.


\subsection*{Curriculum development and synergy} \label{curriculum}

Applied mathematics undergraduate students are required to take a number of courses which fit well with our RTG program; these include MATH 350 Introduction to Computational Mathematics,  MATH 475 Probability, and MATH 476 Statistics. %{\color{red}←\leftarrow please list those courses here; verify this list.}  
%\sout{KEEP?? MATH 350 so far has no coverage of random algorithms. An introduction to Monte Carlo methods will be added to MATH 350 to address this deficiency.} 
Undergraduate trainees will additionally have the opportunity to take additional existing electives %be taking the following existing courses that 
that relate well with this RTG's research Thrusts; for example, MATH 488 Ordinary Differential Equations and Dynamical Systems, and MATH 431 Computational Algebraic Geometry. %, MATH XXX Discrete?{\color{red}Hemanshu? Michael? Rob? Help.}
MATH 431/530 is a dual-delivery (UG/G) course also for graduate students and it has  a heavy project and writing and refereeing component.  Trainees will also enroll in MATH 491 Reading and Research for one or more semesters to obtain credit for research being done during the academic year.  Trainees will be welcomed into our own summer program, but will also be encouraged to spend another summer in an REU or internship elsewhere to gain a broader experience.
%\subsection*{BS Students}  Each year we will recruit 3--4 new undergraduate trainees into our RTG. Beyond calculus, introductory ordinary differential equations, and linear algebra, undergraduates will be strongly urged to take MATH 350 Introduction to Computational Mathematics, MATH 475 Probability, and MATH 488 Ordinary Differential Equations and Dynamical Systems either before entering or early in their traineeships.  These courses will cover important ideas needed in their research.  The first two are required for the applied mathematics major.   PI Duan has taught MATH 488 multiple  times.  He will update the syllabus to integrate recent advances, such as computational and data science techniques for high dimensional dynamics. 
%Add Math 431 which is (Sonja's) course which has a heavy project and writing and refereeing component! writing grading rubrics and all. 
% THIS SEEMS TO BE COVERED BY THE RESEARCH SEMINAR. I'M DELETING IT: 
% By late fall, we will give brief presentations of the research topics of the RTG program, which will serve as an introduction of faculty members of the different groups. Meetings in the  fall and winter will concentrate on training and helping students apply for PhD programs, REUs, and research related internships.  
%\subsection*{PhD Students}
%The majority of our PhD recipients’ first jobs are in non-academic positions. We view our success in placing students in a broad range of organizations as a unique and positive indicator, which we want to maintain. The research topics and interdisciplinary approach should enhance opportunities for jobs in industry and national laboratories for our RTG students and at the same time increase the number of PhD recipients going into high quality postdoctoral positions.
%We will enroll 3--4 new PhD trainees each year, and each trainee will be supported for up to three years, typically their first three year.  While supported by RTG traineeships, these PhD students will be released from other duties to concentrate on their coursework, research, and other RTG activities.  For example, bright students that are lacking some crucial mathematical background may take some upper division undergraduate courses.  

PhD student trainees will complete the foundational courses in analysis, computation, and statistics that are already part of our PhD program.  In addition, they will be urged to take one or more of the following courses which serve the RTG research Thrusts.  
%\marginpar{\bf \color{red} request: new courses 5 lines; old courses 3 lines :)}

%\emph{MATH 544 Stochastic Dynamics}.{\color{red}IS THIS STILL NEEDED?} We will revise this graduate course  incorporating topics on  high dimensional stochastic systems via data-driven algorithms.  The course will fill a gap by addressing simulation methods for nonlocal equations over high dimensional domains.

\emph{MATH 561 Algebraic and Geometric Methods in Statistics}. Algebraic structures are present in a broad variety of statistical contexts, and offer a direct link to the randomized computational tools for Research Thrust 2. The course offers an overview of fundamental theoretical constructions relevant to some of the more popular recent applications in the field. This course first ran Spring 2023 as a graduate elective, with 20 enrolled graduate students. 

\emph{MATH 565 Monte Carlo Methods}. Previously emphasizing mathematical finance, we will revise to include uncertainty quantification. Students will also learn  open source software packages such as SciPy \cite{virtanen2020scipy}, QMCPy \cite{QMCPy2020a} (by Choi, Ding, Hickernell, and collaborators), and DOLFINx \cite{Dolphin12}. 

% \subsubsection*{New Course: Introduction to Mathematical Theories of Complex Fluids} 
% Complex fluids are ubiquitous in our daily life and important in many industrial, physical and biological applications. Studying these materials requires a wide range 
% of tools and techniques for different disciplinary, even within mathematics. A new upper division undergraduate/graduate course will be developed on mathematical theories of complex fluids. The course will be self contained and covers basic mechanics, non-equilibrium thermodynamics, multi-scale modeling and analysis with applications in engineering and biology. The course will fill the gaps between traditional mathematical courses and engineering courses.


%{\color{red}Can this paragraph please be several lines SHORTER? Also "whose course" is this?} {\color{blue} MZ: did} SP: THANK YOU. 
\emph{New Course: Scientific Machine Learning}. 
Machine Learning has become a powerful and popular tool in aiding scientific discovery.  Scientific machine learning, a new branch of machine learning methodology, can achieve improved prediction capability with smaller training data sets than traditional machine learning methods. This course will teach state-of-the-art scientific machine learning methods, including but not limited to Physics Informed Machine Learning, Neural ODEs, Bayesian ODEs, etc.  Moreover, it focuses not only on the theoretical aspects of the machine learning tools, but also on the algorithmic and applications of machine learning.   

 
%\sout{\subsubsection*{New Course: Introduction to Mathematical Framework for Quantum Mechanics} We will build this course from  a   undergraduate course   taught by team member Carlo Segre, and another one taught by our Research Professor Stephen Wiggins (available on YouTube). PI Duan   taught a minicourse in Quantum Dynamics \cite{Gutzwiller1990ChaosIC, Holland1993TheQT,Lindgren2019QuantumMC,Micha2006QuantumDW} in 2019 and 2020 for graduate students and his collaborators.  We will invite our collaborators to co-teach this course.  The YouTube format provides the opportunity to engage the students in an active learning format where they view the “short" lectures ahead of time and the in person class time  also includes  questions, discussions, and problem solving. Stephen Wiggins will co-teach this course with us and his letter of collaboration is a supplementary document.}

\emph{New Course: Analysis of Network Data}, using \cite{KolaczykBook2017} as a main text. Relational data in the form of a network arises in many applications. The course covers the essential statistical inference procedures for analyzing such data, including their shortfalls and open problems that remain to be addressed to properly analyze such data. The computational aspect to the course explores some of the practical questions from  typical applications including  social and information networks. 


\subsection*{Postdoctoral Fellows: Research and Teaching Synergy}
Postdoctoral fellows must become independent researchers with a well-defined research program who can identify interesting, accessible research problems, write research proposals and publications, and teach others.  To hone their teaching skills, postdocs will teach one course per semester. They will help lead the summer programs and research group meetings, prepare and give tutorials on background material, assist in giving minicourses, and help organize workshops.  In these activities they will be mentored by faculty. A detailed postdoctoral mentoring plan is included with this proposal. %{\color{red}QUESTION: does this proposal need to have a postdoctoral mentoring plan as a separate document? probably. do we have that?}

\subsection*{Mini-courses} \label{mini courses}
PIs' collaborators, including SP Leyffer, will offer mini-courses in the form of a series of lectures to supplement regular coursework of interest to RTG trainees. These will be open to all interested students, hence impacting the department more broadly. 
We plan to organize one such mini-course per year, with possible topics including: 
%Our collaborators , at Argonne and other institutions, {\color{magenta}SONJA add Sven's lectures etc.} and some visitors will offer minicourses (each with 10 lecture hours within 1-2 weeks, preferably in summer) to supplement regular courses and to expose trainees to the frontiers of our research projects. Possible minicourses include 
Mixed-integer nonlinear optimization, Nonlinear optimization and machine-learning (Year 1 or 2 by Sven Leyffer), 
{\color{red}OTHER PIs PLEASE EDIT THESE!!$\mapsto$} Machine Learning on Encrypted Data (Year 1),  Non-equilibrium Statistical Mechanics (Year 2),   Bohmian Quantum Dynamics (Year 3), High Dimensional Numerical Analysis (Year 2 or 3), and Foundation of Machine Learning (Year 3 or 4). 
%{\color{magenta}oh geez. "some visitors". :-/} 
%{\color{magenta}Sonja: Sven had said this: I'd be happy to help as Senior Personnell. There are many ways in which we could enhance training from internships to lectures, or courses across the two institutions. I just had the SIAM Student Chapter from IIT visit, and I wll give a careers talk at IMSI later this Spring.  There is also  the network that Mihhail Berezowski has bult at Embry: they pair groups of (undergrad) students with industry projects, see https://mberezovski.com/}

%\FredNote{Do we still want this?  We have a lot going on.} Led by RTG postdocs and graduate students, we will organize RTG  weekly   Colloquia ``Frontiers in Applied Mathematics",  aimed at fostering collaboration between all RTG teams and exposing new research advances to trainees and other members of our department and beyond.   Once a year, the RTG  will organize a Colloquium  under the umbrella of the Illinois Tech SIAM Student Chapter.  



\subsection*{Career Coaching}
Every October, one or two seminar sessions will be dedicated to applying for REUs, graduate schools, and jobs in academia and industry. These sessions will draw upon the experience of our faculty, experienced trainees, and alumni, such as Dr.\ Michael McCourt of SigOpt, a company focusing on Bayesian optimization.  A yearly colloquium presentation  will be offered in September by one of the PI/Co-PIs or experts on writing research proposals. Faculty and postdoc mentors will comment on draft r\'esum\'es,  applications, and proposals.  
The department has organized similar events in the past; namely, 
%We have evidence we have done this in the past:
\marginpar{\color{magenta}TAG: Rob and/or Despina} 
in recent years, SP Ellis and Stasi have worked closely with the SIAM student chapter (???) to organize.... a workshops on job opportunities (both academia and industry?)
%/grant writing for senior graduate students and postdocs;  
We should invite alumnus for homecoming events to discuss their jobs/career path after graduation and such. {\color{red}***we do this...Menger day? please fill in***}






%\subsection*{Workshop on Diversity, Equity and Inclusion (DEI)}
%\sout{Building on our collaboration with Argonne, we plan to co-organize a regularly DEI workshops with the Division of Mathematics and Computer Science Division, whose director, Valerie Taylor, is an African American female, and with Illinois Tech's  Vice President for DEI.  The workshop will consist of two parts: i) encouraging underrepresented groups to pursue STEM careers, and ii) teaching faculty, postdocs, and students how to support and advocate for underrepresented groups.  The frequency will be set so that virtually all trainees will participate at least once.}{\color{magenta}←\leftarrow This seems all too artificial. Kill paragraph.} 

\subsection*{Public Lectures} We will organize a \sout{\color{red}yearly} public lecture in collaboration with  C2ST (see Section 1) to highlight the applications of   our research and excite our audience about the possibilities of mathematics. 
%{\color{magenta} this needs more detials / evidence. budget? letter? at least a draft of a plan?? *Fred* } {\color{magenta}Sonja: why dont' we rely on some other team at IIT already working on this? for the public lecture we say we will organize a satelite event that helps plug in our rtg. if she's willing.? instead of running our own public lecture. OR why don't we just link this to Menger lecture??? MAKE IT DOABLE.}
% At least once during the 5-year window, we will co-sponsor {\bf WITH WHOM?} a distinguished public lecture 
A public lecture on a topic that falls within the Thrusts of the RTG will be co-sponsored  with the department's Menger Day event. \marginpar{\color{red}PLEASE VERIFY THIS IS OK WITH EVERYONE.}
%complex dynamical systems.  The distinguished visitor will come to Chicagoland for a few days, and in addition to the public lecture, also spend time meeting with our RTG faculty and trainees, give a research talk, and share his or her wisdom on being a successful applied mathematician. 
%A letter of collaboration from C2ST is uploaded as a supplementary document.  

%\subsection*{Website and Newsletters} 
%With the assistance of a student worker, we will establish a website for this RTG that serves both an inward facing and an outward facing function.  RTG trainees will help keep the news, events, publications, and bios current.  They will also edit a semi-annual electronic newsletter, featuring our education, research  and outreach  activities. 


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Recruitment and Retention Plan} \label{sec:RandR}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\marginpar{SHORTEN THIS PARAGRAPH, TEXT CAN BE TIGHTER} This RTG involves BS and PhD students and postdocs, collectively referred to as trainees.  
The PIs have a track record of mentoring  applied mathematicians at the undergraduate, graduate, and post-graduate level. 
Our department's faculty have been supervising in-house summer research programs for undergraduate students since the early 2000s; jointly, the PIs and SPs have supervised 16 {\color{red}HOW MANY? Sonja: 12 + Ming: 4 + ... PLEASE FILL IN YOUR NUMBERS!}{
%\color{blue} Ming has supervised 44 UGs at IIT.} **thank you**
undergraduates in summer research projects. 
One faculty member has also supervised a high school student conducting research. 
\marginpar{\color{red}CROSS CHECK numbers.$\rightarrow$ }
During the past five years, our department has trained on average 55 BS students yearly, and these BS graduates have gone to PhD programs in university such as UCLA, NYU, UIUC or found jobs in companies such as Intel, Samsung and Twitter. Our group of RTG PIs has trained 3 PhDs per year (one quarter of them have been female or minority; one third have stayed in academia and two thirds have gone to industry)  and 3 postdocs per year (one third are female; all have gone to academia).  See supplementary Trainee Data and Quantitative Demographic Data. 
%The PIs have  mentored dozensDOZENS????\marginpar{\color{red}CHECK NUMBERS} of PhD students and several postdoctoral researchers, some of whom have entered academic positions at research universities, while others research related jobs in industry and national laboratories. Racial and gender minorities are represented among our current and recent trainees. 
%{\color{red}CHECK: which Li?↦\mapsto}PI Li has mentored an African American female PhD student and an Hispanic PhD student.
%PI Hickernell is mentoring an African American male PhD student.  The PIs have  mentored a number of postdocs, who have been attracted to the vibrancy of our research groups. 
% PI  Pelsmajer, a racial minority, will lead our effort to  recruit African-American, Hispanic, women and other minority students. !!OMG THIS IS TERRIBLE. Anyway, Michael is SP now. 
At present, our department has about \numUG BSs, \numPhD PhD students, and \numPostDoc postdocs. 
{\color{red}Of these, XX \marginpar{Ming?Chun?} are domestic. }


%\subsection*{This needs some work- it is about recruitment. Please edit.} 
%{\color{red}EVERYONE: take a look at this paragraph. Do we still want this? too much going on. Someone please take over leading it [we need a dedicated PI who will do this], or delete.:) it fits well in the recruitment plan.}
%We will create a monthly RTG undergraduate \sout{\color{red}evening - evening???}  event for our applied math majors---both potential and existing trainees.  In early fall, we will give a presentation describing the overall RTG program, with the goal of recruiting students. Subsequent meetings will include presentations from   career services, campus student groups that promote research and innovation, such as the Camras Scholars, Intinium, the Undergraduate Research Journal at IIT which is a formal publication by the Galvin Library's publication arm, %{\color{magenta} which has DOIs! it's a real student publication; our postdocs  and Grad trainees should get involved in editorial work too! and it supports the broader IIT mission - this will be a big plus, i think.}, 
%and also presentations from Argonne and IMSI advertising their programs. 
%{\color{magenta}can also add an "rtg section" to the soremo journal - it has DOIs and all! and is a refereed publication. Gorjana will oversee.} 


\subsection*{BS Students}
Our Department  offers a BS in applied mathematics a variety of specializations as well as options for double majors and co-terminal joint BS-MS degrees. We also offer BS in statistics, and BS and MS in data science.  Some of our majors enter as transfers from junior colleges and some transfer to our major or add our major as a double major after entering Illinois Tech.
 %Each year we will recruit 3--4 new BS applied mathematics or statistics students into the RTG, normally in the middle of their second year standing (as measured by credits obtained).  This may correspond to earlier than their second year at Illinois Tech due to AP or transfer credits.
All majors must take the semester course MATH 100 or Data Science 100, Introduction to the Profession. 
%{\color{magenta}and what is DS100 and should Rob add something?}  
Several course meetings are regularly devoted to faculty research presentations, to which the PIs and SPs will continue to contribute. They serve the dual purpose of introducing general opportunities in the mathematical sciences and connect to potential research opportunities. Our experience shows that this does help recruit students into research; for example, %This works in recruiting students in to research; ofr In the past, these sessions have   We will devote a  couple of sessions research opportunities in the mathematical sciences and introduce our RTG program. \sout{As an aside, 
a Latina first-year student who listened to PI Hickernell's presentation in MATH 100 several years ago did undergraduate research under him as well as under another faculty member, received her BS/MS degree from Illinois Tech and is now pursuing a PhD at the University of Illinois at Chicago.

To attract and retain talented and motivated students, we will provide opportunities for exploring mathematics outside the classroom through our active  SIAM student chapter, who have organized social events such as visiting the Argonne Lab, or a faculty/student seminar series called ``My Favorite Theorem", lunch with a professor, etc. 
We will link our activities with other campus-wide career services, campus student groups that promote research and innovation, such as the Camras Scholars and Intinium. {\color{red}say a phrase about this, what are they? or delete.}
Illinois Tech's library, the Galvin library, has a publication arm which currently publishes the Undergraduate Research Journal  and the SoReMo Journal, both of which formally supported with DOIs. 

%Each year we will co-sponsor with these chapters a session introducing students to our RTG and advising them on how to apply.  The introductory session will feature existing trainees describing their experiences. It may coincide with one of our activities showcasing our research. {\color{red}edit this and incorporate:} Interaction with the SIAM student chapter at IIT?  They have organized social events (like visiting the Argonne Lab, attending CSO etc.), my favorite theorem (a good way for the students to present a talk), lunch with professors, etc.
\sout{MOVE THIS: To increase the diversity and preparedness of the pool of undergraduate students, we will collaborate with our admissions office to reach out to the nearby high schools and junior colleges, which serve a higher proportion of underrepresented minority students.  We will also pitch our RTG to the high school students attending Illinois Tech's summer programs to attract them to apply to Illinois Tech as applied mathematics majors.} 

Each undergraduate trainee will be assigned a faculty mentor, and  an experienced PhD student or postdoc mentor. % This will be on top of our existing departmental advising.  
The mentors will meet individually with each BS student at least once per term to discuss progress, course selection, research plans, summer plans, and career plans.  The mentors will also make themselves available to help the trainees with coursework or research questions. % PI Pelsmajer served for many years as the Associate Chair and Director of Undergraduate Studies. He will coach our mentors on how to fulfill their responsibilities well.

\subsection*{PhD students} 

The proposed activities are well-posed to attract more domestic students, specifically through our proposed collaborative activities with the Argonne National Lab. 
In addition, the proposed co-curricular training activities such as the ``Elements of a research seminar" event will complement introductory sessions at regular research seminars, focusing on filling in the `gaps' students often experience when entering a graduate program, thereby enabling them to continue and succeed when they might otherwise not be able to do so. \marginpar{Hemanshu \& Sonja - Sonja added some text here.}
This RTG will increase the numbers of students and postdocs entering applied mathematics research careers through Illinois Tech and to increase the diversity of our student body.  We will do this by  %strengthening our recruitment and retention efforts and 
 providing enhanced research and communications training, as described in Sections~\ref{section:regular research activities} and \ref{section: co-curricular}.


Our department has a single PhD program in applied mathematics with five emphases: applied analysis, discrete mathematics, computational mathematics, statistics, and stochastics.  Faculty strength covers all five areas and our research projects draw from all of them. Our success in recruiting domestic PhD students is growing, including five new domestic PhD students within the last two years. 
%PhD students act as force multipliers in the research programs of their advisors as well as undergraduate instructional support of the department. 
\marginpar{\color{red}check numbers against budget.} 
To increase the size, quality, and diversity of our PhD program, each year our RTG program will recruit about three new PhD students to be supported for the their first three years of study.  One avenue for increasing the size and diversity of our applicant pool, presently numbering 30--35 per year, will be through our summer REU programs for undergraduates described in Section \ref{sec:summer}.  The SURE program, in particular, involves partnerships with minority serving institutions.  Our summer 2021 SURE pilot engaged an African American student from an HBCU, who is now in our PhD program.  

%Janie Neal, 

The PIs have strong ties to Chicago State University (a predominantly black institution, a Letter of Collaboration is included as a Supplementary Document) and Northeastern Illinois University (an Hispanic serving institution). One African American female was successfully recruited from Chicago State University to our PhD program and nurtured by PI Li.  She is now a faculty member in a public university. A prospective female  African American  PhD student, holding Bill \& Melinda Gates Foundation's Millennium Scholarship, will enter our PhD program in Fall. {\color{red}PLEASE CHECK THIS. is previous one xiaofan's student?}

%Jane has an excellent math and data science background – currently a PhD student at U of Chicago and MS from UIC, in math both.

The PIs will visit minority serving institutions to give seminar talks to students and faculty on mathematical topics of interest related to our RTG.  We will build professional relationships with faculty advisors so that they are confident to recommend their students to our PhD program.  Our PIs will also participate in events that attract underrepresented minorities, such as the annual SACNAS meeting, the AMS joint meeting, and AWM events.

\sout{{\color{red}is this feasible???does it make sense to promise it with this grant?!}Increasing our PhD stipends to make them comparable to those offered by other Chicagoland universities will help recruitment. To do this in a sustainable way, we will partner with Illinois Tech's Advancement Office to raise money for scholarship support for domestic PhD students, while advocating for higher stipends for all of our PhD students.}

%To retain our PhD students, we will assign them 
%Each PhD student will be assigned a faculty mentor from among the PIs when they first enter our program. %  This mentor may change if the student's research interests are best served by a different mentor.  
%The faculty mentor will guide them in developing a portfolio of successful cross-disciplinary research with collaborators outside our department and institution.  This will not only instill in them the ability to bring a variety of mathematical tools to bear on important problems, but also grow their network of research peers beyond their advisor's research group.

%When the PhD student is teaching, the course instructor will observe their teaching, provide feedback.

% The mentor will also be responsible in identifying and recommending the student for suitable summer internships.
%Each PhD student will meet once a semester with the mentor and our Associate Chair for Graduate Studies to monitor progress towards completion of the PhD.  Selection of courses and the thesis committee will be decided in conjunction with the faculty mentor,



%{\color{red}This is about retention... please edit. AND SHORTEN.}
By the end of Year 1, PhD trainees will enroll in one of the groups and be assigned a faculty mentor from the corresponding team. 
The faculty mentor will guide them in developing a portfolio of successful cross-disciplinary research with collaborators outside our department and institution.  %They will also  help  with the summer programs.  % This will not only instill in them the ability to bring a variety of mathematical tools to bear on important problems, but also grow their network of research peers beyond their advisor's research group.
% Year 2 will still be focused on completing necessary coursework.
PhD students will aim to do a research-related internship during the summer of Year 2 or Year 3. Students who do not have
a summer internship will do in-house research and participate in the RTG summer program by giving lectures and mentoring BS students. 
By the end of Year 3, RTG students should complete their coursework, pass qualifying exams, and present their PhD thesis proposal. They should also have participated in one   workshop or presentation run by the Center for Learning Innovation to prepare them for teaching experience. 
In Year 4, students will be fully engaged in their dissertation work. They will also teach a section of Calculus with the support of a teaching mentor, who will observe a few classes and write a report. 
The summer of Year 4 and all of Year 5 should focus on completing their dissertation and applying for jobs and postdoctoral fellowships.
In total, PhD student trainees will have at least two semesters of either supervised teaching or research-related internship. 
{\color{red} !! ↦\mapsto } Internships may be at at Argonne (see letter from SP Leyffer) or other government labs and or at data science companies, such as SigOpt (now a part of Intel) whose Head of Engineering, our alum Michael McCourt, has provided a letter  of collaboration as a supplementary document.\marginpar{\color{magenta}TAG: Chun. verify letters exist with our proposal.} 


\subsection*{Postdoctoral fellows}
We will recruit postdoctoral fellows through our professional networks, approaching promising PhD students that we meet at conferences, and advertisements in \url{mathjobs.org}. Each postdoctoral fellow will have two faculty mentors. One mentor will center on research activities, such as preparation of 
publications, research presentations, mentoring students, and proposal writing. A second faculty member will mentor on teaching activities by way of visits and advising on suitable Center for Learning Innovation workshops they should attend. More details on mentoring are provided in the Postdoctoral Mentoring Plan.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Performance Assessment Plan  }
 
To ensure that we are fulfilling our commitments made in this proposal and making in course adjustments as needed, we will establish and advisory group of three senior academics and administrators, e.g., the Dean of Computing, the Dean of the Graduate College, and the Vice President for DEI. 
{\color{magenta}Hannah said yes! - SP}
The PIs will report to this advisory group on a semi-annual basis and take their advice on how to improve our program.

We will specifically collect and evaluate the following data on the following metrics:

\begin{description}
	\item[Participation] \phantom{a}
	\begin{itemize}
		\item Number of BS and PhD students participating in the monthly colloquia for (potential) trainees
		\item Number and demographics of those applying to and admitted as trainees
		\item Number and demographics of students applying to and admitted into the summer programs
	\end{itemize}
	\item[Outcomes] \phantom{a}
	\begin{itemize}
		\item Destinations (graduate programs, employment) of trainees and summer students
		\item Publications and presentations broken down by demographic
		\item Number of grant proposals submitted and awarded broken down by demographic
	\end{itemize}
	\item[Qualitative Feedback] \phantom{a}
\begin{itemize}
\item Feedback from participants in summer programs, workshops, etc.\ on the success of the program and suggested improvements for the next one
\item Feedback from trainees on the effectiveness of the mentors
\item Feedback from employers/supervisors of program alumni
\end{itemize}
\end{description}
We aim to see greater numbers and a more diverse pool of  applicants to our degree programs than at present.  We also aim to see more students following research careers than at present.  
 


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Organization and Management Plan }
 
 \iffalse
 Provide evidence of past success in training undergraduates, graduate students, and postdocs, identifying names, degree dates, and subsequent placement (including the flow of undergraduates into graduate programs). Applicants may refer to Trainee Data in supplemental documents, if included. Describe the plans, procedures, and personnel for the development and monitoring of all aspects of the project. In particular, discuss plans to ensure appropriate mentoring of students and postdoctoral associates, as well as the roles of the faculty involved. Provide evidence of faculty commitment necessary for the implementation of the proposed program. If the project involves international collaborations, industrial internships, or arrangements with government laboratories, businesses, or other departments, then the proposal should document existing arrangements, any plans for expanding these arrangements, and the personnel involved in managing these linkages.  
 \fi
 
Our university has committed three offices and one seminar room plus an adjacent large open space in the 14th floor of the Michael Paul Galvin Tower for RTG activities. This space will facilitate the training and collaboration that are crucial to our RTG.
 
PI Liu and 4 Co-PIs will oversee the running of the program, as RTG Director  and Co-Directors, respectively.  The tentative rotation has Liu to manage for Year 1, Hickernell for Year 2,    Li for  Year 3, Petrovi\'{c} for Year 4,     and Zhong for Year 5.    PIs, and SPs where appropriate, will participate in the summer activities.
The RTG Director (Liu) will oversee the budget and convene regular meetings of the PIs to ensure that all our activities are running as planned. %Co-PI Pelsmajer has been Associate Chair for teaching for many years and he will be in charge of  training in teaching and mentoring. 
The administrative support will be partially funded by this grant and by the College of Computing. 
For this to work effectively, the Director or Co-Directors will have a course release during the year when they manage the program.


Our trainees will have opportunities for summer interns or other research positions at Argonne National Lab, and SigOpt, an Intel company. %Romit Maulik  is an Argonne scientist with a research faculty appointment at Illinois Tech, and 
 Sven Leyffer (at Argonne) is the research collaborator on Thrust 3.  He will be the liaison person for managing the summer internships at Argonne. %{\color{red}
%We can just add Sven Leyffer here too. 
%Is Romit still on? No, taking it out and replace with the Sven .... Chun} 

 \section{Dissemination}
 We will create a   RTG website that will include a schedule of activities, publications and highlight
distinctions received by RTG participants. The website will link to other items of interest and a link will
be created for recruitment purposes. All presentations by RTG faculty and trainees  will    include a slide advertising our RTG.
Moreover, the resulting publications   will be hosted on arxiv and published in journals.  Software produces will be hosted on GitHub.  Workshop  presentations and  public lectures will be posted on our RTG YouTube channel. {\color{red}please verify: we will indeed have a youtube channel? who will be in charge of this.}

Workshops, public lectures, learning innovation experiences, and research findings will be publicly available on our  RTG website, on our YouTube channel, on GitHub, and in academic journals. We will disseminate our RTG best practices at professional conferences and on our website so that they can inspire national STEM  training programs with partnerships of universities, government labs, industry, and nonprofit organizations.

{\color{cyan}Maybe here we also refer to the mini-courses and all of our seminars which will be open to those outside of RTG --- ??? if there is space -- to tie back in the entire proposal. }



 
%With the assistance of a student worker, we will establish a website for this RTG that serves both an inward facing and an outward facing function.

RTG trainees will help keep the news, events, publications, and bios current.  They will also edit a semi-annual electronic newsletter, 
featuring our education, research  and outreach  activities. 

 
 \section{Post-RTG plan}

This proposal was developed so as to help us institutionalize curriculum and research training changes with the following long-term goals in mind: better prepare applied mathematics graduates for industry and academic careers in an interdisciplinary environment, raise awareness in social responsibility and improve communication training, and increase diversity among our US-based student body. Some of our activities have successfully been piloted in the last couple of years; this RTG will help establish them long-term. Other activities build on our prior collaborative work in education and outreach. The research part of the proposal plays with our small department's strengths and pulls them together in a new way to allow a better synergy between all research activities. 
%
New courses will continue to be offered. PIs have experience developing courses that are in line with the department needs and student demand (see page~\pageref{curriculum}). Inclusion of TT and NTT faculty---who co-direct such programs---in this RTG will ensure overall fit with various undergraduate and graduate programs. 
%We will allow PhD students to co-teach or teach course modules (currently, Illinois Tech does not have student-taught courses).\marginpar{\color{red}VERIFY?}  
The training activities we develop will include materials to be used post-RTG. For example, in-house training materials for communication, mentioned in context of C2ST on page \pageref{C2ST}, will be developed jointly with the Kaplan Institute for Tech Entrepreneurship, through its existing partnership with SoReMo. 
%
With the support of the Illinois Tech administration, we will maintain the postdoctoral program with three fellows each with a 1+1 teaching load. 
We will work with the Office of Advancement to create and sustain special fellowships for underrepresented students. \color{substantiate this claim somehow?? or remove??} 
In summary, we will use the RTG to improve, strengthen, and institutionalize  our ongoing efforts in moving our research and teaching training forward. 
%{\color{magenta} Conclude: and how will we use the success of the rtg to improve the dept going forward!}


% {\color{magenta} --- work in progress --- FOLLOWING IS OLD TEXT: } 
%The infrastructure for recruiting a diverse cohort of undergraduate and graduate students into research groups and supporting their career development through the RTG colloquium will continue and expand throughout our department.  The research groups will actively pursue external research and REU funding opportunities to continue and expand our efforts in the three project areas identified in Section ?????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????????\ref{sec:researchproblems}. The new courses created and the revised courses will continue to be offered.
%The teaching training component for RTG graduate students will allow our department to include undergraduate  classes or teaching modules taught by PhD students. 
%%{\color{magenta}which are currently ZERO. lol} This should be welcomed since all  understand the importance of PhDs having some teaching experience on their resume.
%We will work with the Illinois Tech administration and Advancement Office, as well as through external funding agencies,  to offer both summer stipends and teaching opportunities to all of our post-RTG PhD students.  We will also work to provide special fellowships for underrepresented minority students, as we have begun to do for one such student this fall.
%For an overall assessment of the RTG we will keep contact for the next three years of all participants.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Broader Impacts} 


 

%https://p33chicago.com/

%Although the basic undergraduate and graduate level mathematics curriculum provides sophisticated quantitative tools, it generally does not develop in students the \emph{mathematical maturity} to integrate these tools into a unified \emph{creative toolbox}. Developing this maturity is the overarching goal of this Research Training Grant (RTG).  
 
{\bf Workforce.}
The overarching goal of this RTG is to increase the presence of US citizens, nationals and permanent residents in the workforce of  mathematical sciences. 
By partnering with minority-serving  universities in greater Chicago and beyond, such as   Chicago State University and Northeastern Illinois University, where we already have connections, this RTG   contributes to  the development of a diverse    workforce with technical maturity, interdisciplinary experience, and communication skills, ready to enter academia, industry and government.
%In particular, this RTG simultaneously trains students and postdoctoral fellows to be socially responsible for our environment, health, new energy resources, and financial markets.  They will  better appreciate climate change and rare events, gene mutation and disease, financial risk, energy efficiency, and data privacy in post-quantum era.   
This RTG trainees will further benefit from % contribute to and benefit from collaborative activities with the NSF Institute for Mathematical and Statistical Innovation (IMSI) and Argonne National Laboratory. The trainees will also have 
 opportunities for internships at Argonne, %\marginpar{\color{magenta}we already mentioned internships somewhere else; go back and tighten the text!}
and data science companies such as  SigOpt. The project will  enhance the public’s understanding and appreciation of mathematical sciences and their impact on  society, through our partnerships with C2ST, IMSI, and Argonne.  

This vertically integrated training program will actively recruit women  and racial and ethnic minorities to achieve better equity and inclusion.  The written and oral communication skills of our trainees will be built by requiring them to write to and speak to different audiences, as outlined in Section 2.4. 
In addition to the trainees, the entire Department of Applied Mathematics (both students and faculty) will benefit from this RTG in some way. Students and faculty in the departments of Computer Science, Physics, and Biomedical Engineering will also be positively influenced by the proposed innovative courses %(e.g., Quantum Dynamics, Machine Learning on Encrypted Data) 
and the conducive research environment. 
 

 
{\bf Social impact.} 
The PIs proudly point out that outreach and social impact are pillars of the proposed work in this RTG.
%%\subsection{Outreach} %sonja proposes this: 
%{Outreach and broader engagement}
%{\color{magenta} Outreach is our strength. Sonja will move items to co-curricular etc. then kill this section. Then summarize this and move to braoder impacts section.}
%\subsection*{Socially Responsible Modeling, Computation, and Design
%(SoReMo)}
%{\color{magenta} ***SoReMo does NOT equal outreach.***  SoReMo is directly student training AND outreach. we will also involve phd students and postdocs!} 
 We will incorporate the trainees with the SoReMo projects that are relevant to  social justice and other pressing societal problems. %aspects of social responsibility and could have impact to social justice, including those on the climate dynamics that is crucial in understanding the global warming and environment changes, the biology systems that are related to disease prevention and public health; the engineering applications on battery, fuel cell, energy storage and clean energy; and statistical models connected to social and climate justice. 
%While specific project outcomes are heavily dependent on the projects being carried out, we expect impact on social justice, climate, and other projects as described in the previous section. The following are standard outcomes across all student projects for each and every fellow who is funded by SoReMo: 
In particular, the project will contribute to general  data literacy and engagement with data, whether it be quantitative or qualitative. In fact, SoReMo expands the view of what traditionally constitutes data-driven disciplines. These include: applied mathematics, statistics, computer science, data science, design, social sciences, humanities, architecture. Specifically, we integrate mathematical modeling, computation, data science, the human aspect, and design to make an impact. Sometimes the data is quantitative, sometimes qualitative; gathered through surveys, or  analyzed with statistical tools. Ultimately, though, the data is presented and discussed how it can be used for the common good. 

{\bf Community.} 
This RTG also provides direct impact  on the community and interdisciplinary dialogue at Illinois Tech. 
%Namely, if a SoReMo  project engages in a topic that is related to our campus and community, the outcomes are generally translated to action the semester following the project. 
SoReMo projects impact society by engaging in conversations about difficult issues and proposals of interdisciplinary nature on how to solve them. 
Two of the three spring 2023 projects are directly impacting our community: one studies the effect of redlining to health in the context of environmental justice, and the other studies how to improve outcomes for a particular underserved Chicago community (Englewood) which is trying to be more green in the context of the recycling issues Chicago faces as a city.  %{\color{magenta}can add more on Barien's project as well as Harrison's on the impact on on-campus life, international student onboarding, etc.} 
 
{\bf Inclusion.} Our planned activities are designed to have a positive impact our department  to enhance {inclusion}. 
%\noindent {\bf The department colloquium \& inclusion.} 
% Sonja has edited this part! 
Graduate students and postdocs are expected and encouraged to attend the department-wide research colloquium. The event is already designed so that the invited speakers can interact with graduate students during the after-talk tea session in the department's lounge. In addition, the colloquium coordinator always 
%5) student and postdoc involvement in our department research colloquium, where the colloquium speakers are required to give research talks which are accessible to general audiences; moreover, we also require the speakers to interact with the graduate students for the after-talk tea session in our student lounge.  We arrange 
arranges individual meetings for faculty, postdocs, and students alike, % postdoc and senior graduate students 
to meet with the speakers to discuss their research.  The department has a required graduate-level course for attending the colloquium, which asks students to submit a summaries of talks they attend. 
%, where the students are asked to submit a one-page summary for the talks they attend.  (MZ: this is already being implemented; we don't have a pre-talk-chat-with-students component yet). Yes, it is being implemented but we need the pre-talk chat with students and we need to tell speakers to make their talks student accessible, because most do not. A pre-talk chat if we promise to do it can be something that we do *forever* and therefore can change the culture long-term, and NSF should like that.  
We will enhance this important series of department events by asking our colloquium speakers to do a little bit more work that will leave a large impact: namely, to prepare a short pre-talk-chat with students. Before the formal lecture, the students will meet the speaker to ask questions of ``What is...?" type that we see in the AMS Notices. This will require high-level definitions of concepts that will appear in the talk. While supervised by PIs,  the event will be run \emph{by students} so that the conversation happens on their terms. This is not only more inclusive, but is also more likely to have impact, as previous PI experience suggests.  
% \marginpar{\color{magenta}TAG: Ming} - the research topics represented need to be as broad as we are. It is also the job of the colloquium organizer to **ensure** speakers are student-friendly. Just saying this is not enough; we have to **do** something. I propose creating a 15-minute event before the talk for which the colloquium speaker prepares and which is mandatory for our students: what are the definitions one needs to know in order to follow the talk? What are some intuitive concepts that are critical, but which cannot typically be absorbed by a non-expert in, like, the 3 minutes they show up on a typical slide? Also I do not know what we do for the colloquium credits any more, but bring back the talk summaries (ask Sonja for details) and this is then tying in the entire department into our graduate student training and ensures the impact of the RTG continues post-grant. AND we are all involved. 






 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\section{Results from Prior NSF Support}
%\noindent\underline{{\bf PI Duan:\,}} Duan and Xiaofan Li received prior support from: {\it DMS-1620449, 09/15/2016 - 08/31/2020, \210,000. Theoretical and Numerical Studies of Nonlocal Equations Derived from Stochastic Differential Equations (SDEs) with L\'evy Noises}. Intellectual merit: This project develops efficient numerical techniques for investigating the macroscopic quantities that can help understand the dynamics of SDEs with210,000. Theoretical and Numerical Studies of Nonlocal Equations Derived from Stochastic Differential Equations (SDEs) with L\'evy Noises}. Intellectual merit: This project develops efficient numerical techniques for investigating the macroscopic quantities that can help understand the dynamics of SDEs with \alpha$-stable L\'evy noises, publications include \cite{Chen, WuFuDuan2019, YangDuan2020,ZhangZhuanDuan,WCDKL,CAI2019166,Cai_2017,IJNAM-17-151}. Broader impacts: One African American female Ph.D. student Julienne Kabre, and three other PhD students were supported  by the project. 
 
%%%%%%%%%%%%%%%%% 
\subsection*{Hickernell} NSF grant DMS-1522687, Stable, Efficient, Adaptive Algorithms for Approximation and Integration,
		\$270,000, 8/1/2015 -- 7/31/2018 along with Gregory E.\ Fasshauer (co-PI) and  Sou-Cheng Terrya Choi (female, Senior Personnel).  Other major contributors were Hickernell's research students: 
		six PhDs earned (two female), one current PhD student, and three MSs earned (two female).
Articles, theses, software, and preprints supported in
part by this grant include
\cite{ala_augmented_2017, ChoEtal17a, ChoEtal20a, Din15a, DinHic20a, GilEtal16a, Hic17a, HicJag18b, HicJim16a, HicEtal18a, HicEtal17a, HicKriWoz19a, RatHic19a, GilJim16b, JimHic16a, JohFasHic18a, Li16a, Liu17a, MarEtal18a, mccourt_stable_2017, MCCEtal19a, mishra_hybrid_2018, MisEtal19a, rashidinia_stable_2016, rashidinia_stable_2018, Zha18a, Zha17a, Zho15a, ZhoHic15a}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Intellectual Merit}\marginpar{make it $<5$ lines of text?}
Hickernell and collaborators developed adaptive algorithms for univariate integration, function approximation, and optimization \cite{ChoEtal17a,HicEtal14b, Din15a, Ton14a, Zha18a}.
Hickernell and collaborators developed globally adaptive algorithms for approximating multivariate integrals over the unit cube based on low discrepancy sequences \cite{HicJim16a,HicEtal17a,JimHic16a,RatHic19a}.  The error bounds underlying these adaptive cubatures rely on the Fourier coefficients of the sampled function values, either by tracking their decay rate or by using them to construct Bayesian credible intervals. The cost to bound the error using function data is essentially the same as order the cost to approximate the integral. 
Hickernell and collaborators investigated function approximation problems for Banach spaces defined by series representations \cite{DinHic20a,DinEtal20a}.  For all of these adaptive algorithms, defining a suitable cone of functions to be integrated, approximated, or optimized is key

\subsubsection*{Broader Impacts.}\marginpar{make it $<5$ lines of text?}
Publications arising from this project are reference above.  
The project personnel spoke at many academic conferences and gave colloquium/seminar talks to mathematics and statistics departments and at several conferences.  Hickernell co-organized the 2016 Spring Research Conference and was a program leader for the SAMSI 2017--18 Quasi-Monte Carlo (QMC) Program.   He received the 2016 Joseph F.\ Traub Prize for Achievement in Information-Based Complexity. The algorithms from this research have been implemented in our open source MATLAB Guaranteed Automatic Integration Library (GAIL), which is used in the yearly graduate course in Monte Carlo methods.  Hickernell and the other senior personnel have mentored a number of research students associated with this project, including several female students.  More than a dozen undergraduates have been mentored and several have proceeded to graduate study in the mathematical sciences. 

%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Li} PI Li is supported by NSF grant  DMS-1620449, Theoretical and Numerical Studies of Nonlocal Equations Derived from Stochastic Differential Equations with L\'evy Noise, \$210,000; 09/15/2016 - 08/31/2020. Publications Resulting from this grant:
 \cite{ChenWu, ChenXL2020,  DannyTesfay,GaoTing2016, Gao2016,    Liu2019LvyNI, Lv2016OnAS, QiaoDuan2018,Wang2018NumericalAF, YangDuanWiggins2020,ZhangZhuanDuan,ZhengDuan2017,ZhengYY2020}.

\subsubsection*{Intellectual Merit}\marginpar{make it $<5$ lines of text?}
(i)  \textbf{Numerical Methods}:
    We have developed a convergent and fast algorithm to compute macroscopic quantities from stochastic differential equations  with a-stable L\'evy noises. The numerical results on the mean   exit time, escape probability,  and the probability densities have been obtained \cite{ChenXL2020, GaoTing2016, Gao2016,     Wang2018NumericalAF}.
(ii)  \textbf{Theoretical Analysis}:
We have obtained results on  boundary regularity, interior regularity and maximal principles for   nonlocal partial differential equations for the mean   exit time,   escape probability and Fokker-Planck equation for stochastic dynamical systems with L\'evy noise. This has been inspired also by numerical scheme stability requirement of these nonlocal equations. We have further established a  low-dimensional reduction for a slow-fast data assimilation system with non-Gaussian a-stable L\'evy noise (when only slow variables are observable) via stochastic averaging and slow manifolds \cite{Lv2016OnAS,QiaoDuan2018,   ZhangZhuanDuan,ZhengDuan2017,ZhengYY2020}.\marginpar{\color{red}these simply look like Duan's references. Keep?!}
(iii)   \textbf{Applications}:
   We have revealed certain  dynamical features  for  the evolution of concentration in a genetic regulation model \cite{ChenWu} and a
   FitzHugh–Nagumo   neuronal model,   under additive and multiplicative     L\'evy fluctuations \cite{Liu2019LvyNI}, by examining and computing mean exit time, escape probability, and maximal likely trajectories (based on nonlocal Fokker-Planck equations for the stochastic systems).


 \subsubsection*{Broader Impacts}
 One African American female Ph.D. student, Julienne Kabre, was supported as full-time RA by the project. Three international students were supported during the summer of 2017 by the project.    Another Ph.D. student Senbao Jiang (from Courant Inst-NYU)  has been supported as research assistant under the grant. 
 
 %He has started to develop the numerical algorithm for solving the two-dimensional integro-differential equation induced by the $\alpha$-stable L\'evy noise.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Liu} NSF grants: DMS-1714401,  Topics in Complex Fluids and Biophysiology: the Energetic Variational Approaches, \$349,934, 7/1/2017 - 6/31/2021 and 
{NSF DMS-1950868, Collaborative Research: Multi-Scale Modeling and Numerical Methods for Charge Transport in Ion Channels, \$160,000.00, 8/15/2020 - 7/31/2023.}

 \subsubsection*{Intellectual merit.}  \marginpar{make it $<5$ lines of text?}
% Chun Liu is the PI on the NSF grants:
% DMS-1412005, Energetic Variational Approaches in Complex Fluids and Electrophysiology, \$335,000.00,  8/1/2014 - 7/31/2017.
%grant #2024246 from the United States–Israel Binational Science Foundation (BSF), Israel .
PI Liu's research work during the last funding period was focused on the mathematical problems
arising from the modeling of the elastic complex fluids, together
with their physical, biological and engineering applications.
During that period, the PI had more than 40 peer-reviewed papers published in both mathematics and  other disciplinary journals. A general energetic variational framework had been established for various materials
involving multi-physic and multi-scale phenomenon,  in particular, viscoelasticity and polymers;  ionic fluids and general diffusions.
Some of the accomplishments include: (1) Application of the unified energetic variational framework for a wide class of complex fluids, including ion transport, multi-phase flow,
multi-component flow as well as systems involving moving  surfaces \cite{HuLiLi18,
yang_thermodynamically_2018,benesova_existence_2018, deng_largest_2017,xu_strong_2017,benesova_existence_2018,liu_energetic_2019, Kirshtein2020}, the non-isothermal effects \cite{de2019non,liu2018non, hsieh2020global, Jan-Eric}
and chemical reactions \cite{terebus2018discrete, wang2020field}. 
(2) Papers \cite{liu_energetic_2019, epshteyn2019large, Jan-Eric, hsieh2020global} presenting the analysis of the well-posedness, stability and singularity
for those dynamic systems. (3) Construction of corresponding stable and energy preserving numerical schemes \cite{duan_numerical_2019,xu_numerical_2019,duan_numerical_2019-2, wu2019energetic, liu2020lagrangian, liu2020variational, liu2020structure,duan2020structure}.
 (4)  Application of energetic variational approach to dynamics of ion channel and cell motions \cite{horng_continuum_2019,gavish_bistable_2018, liu_generalized_2017},
and related analytical studies \cite{WaLiTa17, hsieh2020global} with incorporation of thermal effects into these approaches \cite{liu2018non, wu2019energetic, hsieh2020global}.  (5) Development of new coarse grain  methods, including moment closure and projections, to study models involving multiple time scales
\cite{ma_fluctuation-dissipation_2017,ma_coarse-graining_2019}.

% [\color{red} I just noticed that some publication before 2017]}

\subsubsection*{Broader impacts}  Liu worked with
3 postdocs and 3 graduate students, along with several undergraduate students 
on related projects of this proposal. 
% The group also involves  s.
The students learned mathematical, physical and biological
aspects of the theory, while actively collaborating  with researchers in those fields.
They also attended and presented their work in workshops and conferences.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection*{Petrovi\'c} 
DMS\#1522662, Collaborative Research: Randomized and Structure-Based Algorithms in Commutative Algebra, 8/25/15-8/25/19, \$229,577. (Total collaborative award \$430,000.)
\\
\emph{Intellectual merit.} The project offered  a  significant speed-up, from exponential  to expected linear runtimes, to   solving systems of polynomial equations %and finding ideal generators   
using a probabilistic method.  The increase of use of combinatorial structures in polynomial computational problems can be felt across applications including statistics, optimization, graph theory, and systems biology. 
 \\
\emph{Broader impacts.} This grant was used to train 3 Ph.D students who wrote dissertations around the topic, as well as train and supervise several undergraduate students. PIs wrote and published software to be used by other researchers. It funded a  summer school at Max Planck Institute in Leipzig, Germany,  with $70+$ participants. 

\\
\emph{Products.}
 The project has resulted in 18 papers: \cite{GKP-fienberg, RMI, semigroups, AlgStatNtwk-ARSIA, WhatIsMB, jalgstat-fienberg-editorial, karwa2016dergms, hypergraphs-toric, jesus1, jesus2, jesus3, jesus4, jesus5, jesus6, jesus7, algebraicViolators, bouquets, surveyAlgStatCONMchp},   
 and a software package \cite{RMIm2}. 

PI Petrovi\'c  lead the organization of a focused 
 conference on Algebraic Statistics, co-sponsored by DMS \# 2004271 (co-PI), ``Algebraic Statistics 2020", in the amount of  \$34,000. 
\\
\emph{Intellectual merit.} Being the third large US-based meeting focused solely on the field of algebraic statistics,  the conference has provided a focused opportunity for interdisciplinary collaboration. 
\\
\emph{Broader impacts.} 
The conference, Algebraic Statistics 2022, was held at the University of Hawaii at Manoa in Honolulu, HI from May 16, 2022 to May 20, 2022. The conference had 10 keynote talks, 21 shorter 20-minute contributed talks, and 8 posters, most were early-career participants. %These talks introduced the audience to new work in algebraic statistics, ranging from algebraic methods for graphical models, sparse polynomial system solving, establishing parameter identifiability for models from evolutionary biology, geometry of polynomial optimization, and geometric principles of machine learning and statistical inference.
%The conference had 53 registered participants from 8 countries, representing % with the following fields represented (field determined by department of academic appointment): 
%Mathematics, Statistics, Applied and Computational Math, Computer Science, Engineering, Operations Research, and Computational Biology departments.  
%In addition to the conference, we held two complementary activities: a graduate education luncheon to discuss and share best practices in mathematics graduate education and a week-long workshop aimed at junior researchers.

\noindent 
\emph{Products.} For this conference, titles and abstracts have been posted on the  AS2022 conference website: https://algebraicstatistics.org/as2022.
%The organizers of the conference invited submissions to the special issue of Algebraic Statistics on the occasion of the conference.
%Journal Editors: Thomas Kahle and Sonja Petrović, 
%Special issue co-editor: Elizabeth Gross. 
There will be a volume in Algebraic Statistics (an MSP journal) to be published in 2023 containing  more than a dozen of these papers. 

PI Petrovi\'c is currently supported by DOE Office of Science, Randomized Algorithms for Scientific Computing, Award number 1010629. This is joint with Argonne National Laboratory (2022-2025) in the amount of \$2.8M.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection*{Zhong} PI Zhong is supported by NSF grant CCF-AF-$2225507$, A Bayesian Paradigm for Physics-Informed Machine Learning, \$$586,319.00$, $12/1/2022$ - $11/30/2025$.  Publications resulting from this grant:~\cite{chen2022using, coutinho2022physicsinformed, ZLABN2023}.

\subsubsection*{Intellectual merit.} 
We plan to develop state of the art and novel Bayesian based scientific machine learning methods to solve PDE-related forward and backward problems with applications in petroleum engineering, mechanical/aerospace engineering, astrophysics/astronomy. 
%\marginpar{make it $<5$ lines of text?}
\subsubsection*{Broader impacts} 
The grant will support undergraduate and graduate students as well as postdoctoral researchers.  It will create a new graduate level course in scientific machine learning at both Texas A$\&$M and Illinois Tech. It will foster a community of faculty members, students, scientists within both Texas A$\&$M University and Illinois Tech for scientific machine learning for organizing seminars, workshops, webinars, and summer schools.  

% \medskip 
 
%\noindent  \textbf{}: No current NSF funding or   awards with an end data within the past five years. 



%Therefore, there are recent   machine learning methods devoted to discovering governing laws of nonlinear phenomena from noisy data. Some of these works are based on Kramers-Moyal formula \cite{DaiMinChaos} or Koopman generator  \cite{WuFuDuan2019, LuYB2020, Yanxia2020}. 


\newpage
\pagenumbering{arabic}
\renewcommand{\thepage} {\arabic{page}}

\bibliographystyle{abbrv}
% \bibliographystyle{plain}
%\bibliographystyle{natbib}
\bibliography{asi,ref_art,stochas,ml_ref,datasci,VD, quantum,FJHown23,FJH23,LDS,NSF_refs,productsCompAlgNSFdeloerapetrovicstasi,MarkovBases,generalNetworks,AlgStatAndNtwks}

\end{document}    % End %







%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Material Deleted from above is given below}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\iffalse  
 While there   may be thematic driven differences on the approach to deliver knowledge and do research, each group share a similar year-round schedule that will allow awareness
and exchange of ideas throughout the academic year, mainly at the RTG graduate seminar.   Finally each summer program is centered on
an intense 2-week period. \FredNote{What does this mean?} Participating undergraduates will have a unique opportunity to work in a serious way on two topics, which we believe to be highly beneficial.
\fi


% Talk about this later:  In fact, IMSI has encouraged us to propose a joint  workshop relevant to our RTG Thrust,   thus helping publicize and recruit students and postdocs for this RTG.

%Other entities at Illinois Tech relevant to this RTG include the  Center for Interdisciplinary Scientific Computation,   Laboratory for Stochastic Dynamics and Computation, Center for Learning Innovation, and Career Services. The Camras Scholars Program help recruits and retain talented undergraduate students. Our Interprofessional Projects (IPRO) Program provides the alternative to a traditional undergraduate education. Our signature IPRO Program remains one of just a few programs of its kind in the country. IPRO joins students from various majors to work together to solve real-world problems, often on behalf of sponsor companies and nonprofits. A required academic program, IPRO teaches leadership, creativity, teamwork, design thinking, and project management—uniquely preparing students to succeed in a professional work environment. Moreover, the  Pritzker Institute of Biomedical Science and Engineering provides a limited number of research stipends for undergraduate math majors to conduct research in biophysical modeling. 
   
%All of our 18 tenure-stream faculty are actively engaged in research and teaching. 

%All faculty are fully engaged in teaching both undergraduate and graduate studies.   Current faculty members (Names?) have received the highest teaching recognition:.... Distinguished Teaching Professor Award. The PI Duan once received a Teaching Innovation Award. We also have four postdocs or visiting assistant professors with significant teaching responsibilities. 

%With this in mind, it is not unrealistic to expect that in one form or another, all Math faculty will participate in this RTG program, either as research team co-leaders, participants, or teaching mentors. In the next section and in the budget, we will identify those colleagues who will carry out specific RTG related tasks.

%We are in an ideal location    for research  training activities. For example, we hosted 3 CBMS-NSF research conferences and attracted .... 


\iffalse 
state-of-the-art deep learning architectures is essential for multiplying opportunities for applied
mathematics undergraduate and graduate students in data-driven modeling and simulation. Thus
this project trains students and postdocs, not only in their professional skills, but also in social
responsibility, through better understanding of climate change, gene regulation or mutation, and
other processes subject to abrupt transitions.


Quantifying transition phenomena. To understand transition in a stochastic system, we study the
‘ensemble’ (or ‘continua’) quantities: mean exit time (exiting from a dynamical regime), escape
probability (from one dynamical regime to another), and the transition probability density. These
provide tools to better understand climate shifts [169] and gene transcriptions [16]. This leads to
deterministic nonlocal (integro-differential) equations for mean exit time and for escape probability,
and a nonlocal Fokker-Planck equation [36], in which the usual Laplace operator is replaced by a
nonlocal Laplace operator (which is related to the generator of L ́evy motion Lα
t ): −(−∆) α
2 , for
α ∈ (0, 2).
As the usual numerical methods (such as finite difference) are not feasible, we will develop a
method to simulate these high dimensional nonlocal equations based on a stationary or temporal
normalizing flow model, as a generative machine learning tool. This method involves techniques
to handle the nonlocal terms implicitly since arbitrary probability densities can be learned from
samples alone. Subsequently, ‘ensemble’ quantities, including the mean trajectories (when exist),
are high dimensional integrals with respect to the probability density [36]. Thus we will validate
our fit distributions with Monte Carlo simulations, in particular highly stratified or quasi-Monte
Carlo variants, where our team has made recent theoretical advances [52, 56] and efficient imple-
mentations [20].
Furthermore, we will investigate the most probable transition pathway between dynamically
different regimes (e.g., metastable states) during finite transition time. This pathway is the minimizer of of the associated Onsager-Machlup action functional, $A(z,  ̇z) =\int_0^T L(z(t), ̇\dot z(t))$dt, where
the Lagrangian L depends on the vector field f , noise intensities g, σ, and jump measure ν for
L\'evy motion; for background, see our recent works [13, 64]. The transition time T may also be
estimated [63]. We will explore how the energetic variational approaches in Project 2 may benefit
us in the minimization of the Onsager-Machlup action functional here.
Extracting transition phenomena from noisy data. Some stochastic systems may only be known
or partially known (where some mechanisms are missing) by noisy observations or simulated data
sets. We will develop techniques to extract stochastic governing laws [79, 80, 101] and then examine
transition phenomena by computing mean exit time, escape probability and transition probability
density as above. There are several potential interfaces with state-of-the-art generative machine
learning here which has been constructing novel methods to learn probability densities from data
as neural stochastic differential equations. Furthermore, our colleagues’ works on designing and
optimizing sensor networks, and graph theory [73, 11] will be explored to guide smart data collection
for interface with such novel algorithms.
A class of high dimensional dynamic systems may be interpreted as networks, such as gene reg-
ulation networks [123] which are modeled by systems of stochastic differential equations [140]. In
a gene regulation network, a transition from low to high concentration of a promoter protein may
indicate transcription [138, 168]. Studying transitions in large networks is also of vital importance
for resiliency analyses and control of powergrid infrastructure and for supply chain modeling in
operations research which may be modeled as stochastic processes on graphs [135, 4, 115]. For
large networks with multiple sources, sinks and connectivity measures of several variables, ex-
tremely high dimensional systems are commonly obtained. Furthermore, optimization of these
graphs for desired quantities of interest require the solution of a complex mixed-integer nonlin-
ear programming problem with constraints that may not be nominal input-output mappings but
aforementioned complex dynamical systems [136, 132, 74, 134].
Project 2. Energetic Variational Approaches for Complex Dynamics. Chun Liu (leader),
Yiwei Wang (Illinois Tech, Applied Math), and Georgia Papavasiliou (Illinois Tech, Biomedical
Engineering).
Background. The research focus will be on the applications of the energetic variational approach in
mathematical modeling, nonlinear partial differential equation, scientific computing and machine
learning. The framework of energetic variational approach, originated from seminal works of
Raleigh [139] and Onsager [119, 120], provides a paradigm to determine the dynamics of a non-
equilibrium system from prescribed energy-dissipation laws. These approaches have motivated
much work in the area where it is crucial to preserve the underlying thermodynamics structures,
including quantification of noises and fluctuation of the systems. The variational principle has
been employed to study many complex fluids, such as liquid crystals, two-phase flows, and ionic
fluids, which involve the coupling and competition of various mechanical and chemical mechanisms
in different scales [82, 41, 83, 84, 87, 32, 141, 39, 48, 92, 76].
In an isothermal closed system, an energy-dissipation law, coming from the first and second law
of thermodynamics, is often given by d
dt Etotal = −4, where Etotal is the total energy, including
both the kinetic energy K and the Helmholtz free energy F, and 4 ≥ 0 is the rate of the energy
dissipation which is equal to the entropy production in the situation. The energy-dissipation law,
along with the kinematics relation, capture all specific physical and biological properties for a
given non-equilibrium system. Starting with an energy-dissipation law, the EnVarA framework
derives the dynamics of the systems through two variational principles, the Least Action Principle
(LAP) and the Maximum Dissipation Principle (MDP). The LAP, which states the equation of
motion for a Hamiltonian system can be derived from the variation of the action functional A =∫ T
0 [K − F]dt with respect to the flow maps, gives a unique procedure to derive the conservative
force for the system. The LAP is indeed a manifestation of the rule δE = force · δx. The MDP,
variation of the dissipation potential D, which equals to 1
2 4 in the linear response regime, with
respect to the rate (such as velocity), gives the dissipation force for the system. In turn, the
force balance condition leads to the evolution equation to the system, δD
δxt = δA
δx . Through different
choices of the free energy F and the dissipation 4, the EnVarA systematically deals with the
coupling and competition of mechanical, chemical and thermal mechanisms in different time scales
in a thermodynamically consistent framework. It provides an unified framework to study various
dynamical systems demonstrated in the other projects, and moreover, gives a way to quantify
energy transduction between different processes.
Applications in Biology and Engineering. The team has the track record and relevant expertise to
lead the projects. We plan to focus on projects related to the modeling of active soft materials in
biological environments. The unified framework will provide the insights for the systematical study
of coupling and competition of these multiscale and multiphysics systems. The interdisciplinary
nature of the team will provide an opportunity to have balanced training in different approaches
of applied math, most notably multiscale modeling and analysis associated with the applications
in biology and soft matter physics.
Physics Preserving Numerical Simulations. The energetic variational formulation also provides
guidelines to develop numerical schemes that preserves the basic physics, such as frame-indifference,
positivity of physical quantities and conservation of mass, for many complicated systems arising
from multiscale multiphysics problem in physics, biology and material science [91, 90, 89]. The
same numerical frame can also be applied develop new algorithms for machine learning [151].
Different spatial discretizations, including finite element, spectral methods, particle methods and
neural network discretization, and time-stepping techniques can be incorporated in the numerical
framework. These projects will provides training in the areas of scientific computation, multi-scale
simulation and machine learning.
Project 3. Mathematical Tools for Open Quantum Systems. Jinqiao Duan (co-leader),
Fred Hickernell, Xiaofan Li, Carlo Segre (co-leader, Illinois Tech Physics), Romit Maulik (Argonne
National Lab \& Illinois Tech) and Martin Suchara (Argonne National Lab).
Background. The development of quantum science and its exploitation in technology is a topic of
great interest and activity in many countries. The United States [126], Europe [127], and China
[72] have invested heavily in long term “quantum initiatives”. It has been stated that we are in the
midst of the “second quantum revolution” [72]. Indeed, most people tend not to realize just how
much quantum mechanics influences their day-to-day life. For example, consider that standard
household light bulb. Advances in light emitting diode (LED) technology have led to “white” light
bulbs that can now be purchased anywhere light bulbs are purchased. These LED light bulbs
are more energy efficient, last longer, and are more environmentally friendly than light bulbs of
the past. This is all thanks to quantum mechanics, which is fundamental to LED physics and
technology. The impact of this rather mundane example is significant. But the impact of more
complex physical and technological quantum advances, such as quantum computing, quantum
cryptography, quantum sensors [116], and even quantum artificial intelligence [142] will change the
world and our way of life.
Therefore it is essential in education nowadays that students develop a level of “quantum liter-
acy” [42] that will enable them to interpret and evaluate these new phenomena and technologies in
a way that will enable them to assess their impact, not only on their own lives, but on society and
the world. Developing this quantum literacy in the context of mathematical education, at both
the undergraduate and graduate levels, is a goal of this RTG project.
Quantum mechanics is not typically emphasized in most mathematics programs. Indeed, many
of the unusual properties of quantum mechanics, such as tunneling, entanglement, superposition
of multiple states, interference of states, uncertainty, and quantum measurement can be under-
stood from the fundamental physical principle of “wave-particle” duality and de Broglie’s relation
between the momentum and the wavelength of a particle. From this one is led to the Schr ̈odinger
equation description of particles in terms of waves. In fact, the physical setting of quantum
mechanics is described by a few (mathematical) axioms that describe the state space, observ-
able quantities, measurement, and time evolution. Once this physical setting is translated into
mathematics, the implications and consequences of the mathematics is a natural setting for the
participation of mathematicians. However, the required mathematical skills and knowledge are
not typically emphasized or even covered in the relevant courses. A goal of this RTG is to remedy
this situation, both through training undergraduate, graduate, and postdoctoral students, and
through the development of a course appropriate for mathematicians, but which bridges this gap.
This course will lead naturally into a variety of research areas that fit naturally into the context
of the other research topics, that we will describe. Indeed, we will explore how Project 3 helps us
in machine learning of encrypted data, especially for post-quantum era, linking with Project 1.
Exploring the Interplay Between Quantum Mechanical Uncertainty and Noise. Environmental
noise arises naturally in the mathematical formulation of open quantum systems [81, 10, 105, 50].
This project will explore the interplay between quantum mechanical uncertainty and the influence
of external noise [86, 114] on the quantum mechanical system, including metastable transitions
[105, 128] and the connection with our earlier works on the most probable transition pathways
in stochastic systems with non-Gaussian L ́evy noise [13]. Moreover, we will examine how certain
\fi 

  

\iffalse 
\subsection*{Illinois Tech resources  for this RTG}
Our university has the following resources to support this RTG.

 $\bullet$ Center for Interdisciplinary Scientific Computation
 
 $\bullet$ Laboratory for Stochastic Dynamics and Computation
 
 $\bullet$ Pritzker Institute of Biomedical Science and Engineering
 
  $\bullet$  The Camras Scholars Program
  
  This Program help recruits and retain talented undergraduate students. 
  
  $\bullet$  Center for Learning Innovation 
 
 
 $\bullet$   Career Services
 \fi
 
% Argonne National Laboratory resources ....
 
 
% 1. Introduce with connections for stochastic dynamical systems and deep learning and add literature review. 
% 2. Justify for RTG.


% Recent literature in deep learning have proposed the development of novel architectures that interpret the various hidden layers of feedforward neural networks as the discrete time steps of an ordinary differential equation. Neural ordinary differential equations \cite{chen2018neural,rubanova2019latent}   consider the continuum limit of neural networks where the input-output mapping is realized by solving a system of ordinary differential equations.
% %\begin{equation*}
% %    \frac{d \mathbf{u}}{d t} = f(\mathbf{u},t,\theta), \quad %\mathbf{u}(t_0) = \mathbf{u}_0, \quad t \in [t_0, t_F]
% %    \label{eq:node}
% %\end{equation*}
% %where $\mathbf{u}_0 \in \mathbb{R}^D$ is a feature vector for the input data, $\theta \in \mathbb{R}^N_p$ are the weights of the neural ODEs, $f: \mathbb{R}^D \times \mathbb{R} \times \mathbb{R}^N_p  \rightarrow \mathbb{R}^D$ is a neural network.
% %Essentially, the input-output mapping is learned by approximating %the vector field $f(u,t)$ through a data-driven approach. 
% This continuous model makes neural ODEs particularly attractive for learning and modelling the nonlinear dynamics of complex systems. They have been successfully incorporated into many data-driven models and recent literature has delved deeper into the connection between neural architectures and stochastic dynamical systems \cite{}. Therefore, the study of data science problems that can be framed as the solution of complex dynamical systems is essential for multiplying opportunities for applied mathematics undergraduate and graduate students in data-driven modeling and simulation. 
 
% and Riemannian geometry \cite{lou2020neural,falorsi2020neural} 

%For instance, it has been argued that diffusion by geophysical turbulence \cite{Shlesinger} corresponds  to a series of  ``pauses", when the particle is trapped by a coherent structure, and ``flights" or ``jumps" or other extreme events, when the particle moves in the jet flow. Paleoclimatic data \cite{Dit} also indicate such irregular processes. There are also experimental demonstrations of L\'evy flights  in optimal foraging theory and rapid geographical spread of emergent infectious disease.   Humphries {\it et. al.} \cite{Humphries}   used GPS to track the wandering black bowed albatrosses around an Island in Southern Indian Ocean to study the movement patterns of searching food.   They found that by fitting the data of the movement steps, the movement patterns obeys the power-law property with power parameter $\alpha=1.25$. To get the data set of human mobility that covers all length scales,  Brockmann   \cite{Brockmann}  collected data by online bill trackers, which   give successive spatial-temporal trajectory with a very high resolution. When fitting the data of probability of bill traveling at certain distances within a short period of time (less than one week), they found power-law distribution property with power parameter $\alpha=1.6$.

%This parameter is of great importance since by using it into the classic SIS model, they found probability density function patterns generated by non-local dispersal:  $\alpha-$stable L\'evy motions are strikingly similar to practical data of human influenza. A further example is a thermally activated motion of a test particle along a polymer,   shown in \cite{sokolov1997paradoxal, brockmann2002levy} to be subject to $\frac{1}{2}$-stable L\'evy motion due to polymer's self-intersections.  This motivates the investigation of dynamical systems driven by both Gaussian and non-Gaussian   fluctuations, especially the heavy-tailed,  $\alpha-$stable   L\'evy motions.  

%Both mean exit time and escape probability are described by nonlocal `elliptic' partial differential equations, with unusual boundary conditions corresponding to absorbing or escaping situations at microscopic levels. The probability density function for the solution of an SDE with  with L\'evy noise is described by a nonlocal Fokker-Planck equation.
 
%between distinguished states \cite{DaiMinChaos,LuYB2020, ZhengDuan2017, HuangYF, ZhengYY2020}.  and the Onsager-Machlup action functionals \cite{ChaoDuanOM, HuangYF2020}.  We will combine our group member' research in networks and    graph theory,  to reveal dynamical transitions between distinguished states (e.g., metastable states) of these large networks.

%Machine learning for high dimensional stochastic dynamics using deterministic quantities or indexes which are defined as mean or high dimensional integrals.

\iffalse
\noindent \textbf{Regular group activities:} We will hold   the weekly Stochastic Dynamics Seminar. The key component     is for     students present articles relevant for the main topics of the
project followed by a discussion of how the methods developed in our research could be applied to emerging
areas of   research (e.g., gene networks, climate models, machine learning inspired by stochastic dynamics). This will enhance our students’ grasp of the important current research topics
and potentially lead to new collaborations. We also propose that graduate students and postdoc regularly
make short presentations on their research progress to all RTG participants, followed by a discussion of next steps in each individual research project.


 \textbf{Yearly workshop:} A two-day workshop will be organized in the end of each academic year. Each workshop will consist   of presentations by graduate
students and postdocs, with one or two invited lectures (funded internally by the math department) given
by top researchers. These researchers will be chosen from experts.   
\fi

%Noise The numerical scheme .. can capture the coupling and the compeation, Besides,   the algorithm can also have applicaion in machine learning data science 


\iffalse
1. General energetic variational framework for complex fluids: 
      a) least action principle and maximum dissipation principle. 
      b) Navier-Stokes equations and elasticity. 
      c) viscoelastic materials: nonlinear elasticity, incompressible elasticity, viscoelasticity. 
      d) generalized diffusion, nonlocal diffusion.

2. Free interface motion in the mixture of different fluids: 
     a) conventional description: sharp interface description, water wave, vortex sheet, surface tensions. 
     b) diffusive interface description: microscopic background (self-consistent field theory), Flory-Higgins theory, 
         sharp interface limit, dynamics. 
     c) slippery boundary conditions and systems on (or near) surfaces. 
     d) Helfrich elastic bending energy and application to vesicle membranes. 

3. Multiscale modeling and analysis: 
     a) basis of stochastic differential equations: Fokker-Planck equations, diffusion, 
         Smoluchowski coagulation equations, variational formulations, kinetic theory. 
     b) micro-macro models for polymeric materials. 
     c) moment closure methods, Mori-Zwanzig formulation and other coarse grain methods. 

4. Ionic fluids and ion channels: 
     a) electroeheological (ER) fluids: Poisson-Boltzman fluids, like charge attaction (LCA) and charge inversion, 
         steric effects of ion particles, density function theory of Rosenfeld, equation of states. 
     b) basic physiology of ion channels and protein structures. 
     c) ionic fluids in ion channels. 
     d) ionic osmosis in biological systems. 

\noindent {\bf Regular group activities}: {We will continue the weekly Multiscale and Complex Fluids Seminar. One key activity is to have students present articles relevant for the main topics of the project and explore the potential applications. This could also foster new collaborations between the students and other researchers. We also propose that graduate students and postdoc regularly make short presentations on their research progress and discuss the possible new approaches.}

\noindent {\bf Yearly workshop}:  A two-day workshop will be organized either in Illinois Tech. We will also pursue opportunities for collaboration with international centers such as Banff International Research Station and the NSF Institute for Mathematical and Statistical Innovation (IMSI).
\fi


%\smallskip
%\noindent{\bf Simple Models Describing the Phase Space Formulation of Quantum Mechanics.} The potential well problems in the first two projects are formulated and solved in the standard Schr\"{o}dinger equation setting.  This project will explore reformulating those problems in the phase space setting using the Wigner function approach. The two approaches will be compared and contrasted.

%This is related to the ‘’Lagrangian transport’’ project.

%\smallskip

%\noindent
%{\bf Explorations of the Role of Bifurcations of Equilibria in Hamiltonian Systems on Classical Reaction Dynamics.} For two degree-of-freedom Hamiltonian systems the saddle-node, pitchfork, and Hamiltonian-Hopf  (or the (‘’Meyer-Schmidt’’) bifurcations typically occur.  We will explore how the normal forms for these bifurcations occur in chemical reaction dynamics and their influence on reaction.
  
%\item[Explorations of the Role of Bifurcations of Equilibria in Hamiltonian Systems on Quantum Reaction Dynamics.] This project will consider the classical normal forms of the previous project and the quantum mechanical manifestations of these bifurcations in reaction dynamics.

\smallskip

%Our training and research in this project will use the open source software QuTiP \url{http://qutip.org} and Qiskit \url{https://qiskit.org}, which are now standard tools for modelling and simulation of quantum phenomena.

\iffalse 

\noindent \textbf{Regular group activities:} We will hold   the weekly Quantum Dynamics Seminar.      Trainees in this team will present articles relevant for the main topics of the
project followed by a discussion of    next steps in   individual research issues.


 

\noindent \textbf{Yearly workshop:} A two-day workshop will be organized in the end of each academic year.  It will consist   of presentations by graduate
students and postdocs, with several invited lectures 
by top researchers.  
\fi

\iffalse
Our Interprofessional Projects (IPRO) Program provides the alternative to a traditional undergraduate education. Our signature IPRO Program remains one of just a few programs of its kind in the country. IPRO joins students from various majors to work together to solve real-world problems, often on behalf of sponsor companies and nonprofits. A required academic program, IPRO teaches leadership, creativity, teamwork, design thinking, and project management, uniquely preparing students to succeed in a professional work environment. Our department offers a recurring IPRO equivalent (an ``I Course'', taught as MATH 497) which has a special focus on techniques from mathematics and statistics.  Moreover, the  Pritzker Institute of Biomedical Science and Engineering provides a limited number of research stipends for undergraduate math majors to conduct research in biophysical modeling. 
\fi



Throughout the academic year, there
will be research presentations by RTG participants, mostly graduate students and postdoctoral fellows, with
more emphasis during the spring term. This last component serves two main purposes: to expose undergraduate
students to research and to prepare students for the four-week RTG summer program. The main
objective of this activity will be to increase the number of math and statistics majors that apply to PhD programs.
RTG funds will be used to provide food for these events, materials for presentations and advertising.



%\smallskip
%\noindent
%{\bf Linear Algebra from the Quantum Point of View.} It cannot be emphasized enough that Linear Algebra (Linear Space Theory in general) is fundamental to many areas of pure, applied, and computational mathematics, and it is also fundamental to the structure of quantum mechanics. Of course, ‘’every’’ mathematics department has courses in Linear Algebra. However, topics important to quantum mechanics are often omitted.  Our course will fill that gap. The starting  point will be to emphasize that the basic structure of quantum mechanics is complex in nature—the solution of Schr\"{o}dinger’s equation is complex valued and linear spaces are complex vector spaces. The language that we will use is Dirac notation.  This complex structure has implications for the nature of inner products, and this plays a key role in the mathematical structure of quantum mechanics.  Multi-particle quantum systems are essential for the notion of entanglement. In classical mechanics multi-particle systems are described by the cartesian product of state spaces for each particle. In quantum mechanics they are described by a tensor product, and this algebraic structure tends not to be treated in traditional linear algebra courses, but it is essential for describing the basic elements, and showing the advantages of, quantum computation, and this mathematical structure will be emphasized in our course.  The Schmidt decomposition provides an insightful characterization of entanglement based on tensor product spaces and its proof relies on the singular value decomposition, another topic of great importance (also in data science) that is typically neglected in a first linear algebra course, but it will be emphasized in our course.




%We envision one lecture to focus on developing models for physical systems and the other lecture to be primarily addressing the basic concepts behind the numerical methods used for simulations of such systems. For the rest of the two weeks, the focus will be on three sets of activities: (i) seminar-style presentations prepared by graduate students and postdoc, describing specific topics related to the ongoing research projects; (ii) undergraduate students working in teams on small research projects chosen in the beginning in the first week under the guidance of RTG faculty; (iii) visiting   Argonne National Laboratory and Pritzker Institute for Biomedical Science and Engineering and interacting with the faculty members, students, and postdocs conducting experimental works, with the objective of getting better understanding of the physical systems being modeled in the framework of the project. Each team of the undergraduate students will make a short presentation of the results of their work in the end of the second week.
 
% and minicourses on Data Science for Dynamical Systems, Advances in Multiscale Modeling, and Introduction to Quantum Dynamics offered by distinguished RTG visitors. 
 





%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%% Material Deleted from above is given above
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%













%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Supplementary Documentation}
 

\subsection{a. Letters of Collaboration}

--- from Argonne Lab


Signed letters of collaboration by the institution and other sources in support of the project should be included. If industrial or government laboratory internships are planned, letters indicating the willingness of the external organization and of individual external mentors (if known) to participate should also be included. These documents should be scanned and uploaded into the supplementary documentation section.
The letters of collaboration are meant to explain how the institution and the collaborating sites will provide an environment that supports the proposed research and training activities. It is acceptable for a letter of collaboration to briefly mention specific activities supported by the collaboration and listed in the proposal; however, each letter is limited to one page. Letters of recommendation or endorsement are not allowed.

\subsection{b. Trainee Data} 

All applicants are strongly encouraged to supply the following data. Note that data is requested for the group submitting the proposal, not for the entire department. For new RTG proposals, data should be included for the past five years. For a renewal of an existing RTG grant, data should be included for the past ten years.
A list of Ph.D. recipients, along with each individual's baccalaureate institution, time-to-degree, post-Ph.D. placement, and thesis advisor.
A list of postdoctoral associates (including holders of named instructorships and 2- or 3-year terminal assistant professors), their Ph.D. institutions, postdoctoral mentors, and post-appointment placements.
The dollar amount of funding by federal agencies for Research Experiences for Undergraduates (REUs), graduate students, and postdoctoral associates.





\subsubsection{A list of graduate students}

Romit:

Summer graduate students through Argonne:
Suraj Pawar, Oklahoma State University, Research Aide, 2020
Dominic Skinner, MIT, NSF MSGI, 2020
Alec Linot, University of Wisconsin Madison, Wallace Givens Associate, 2021
Sahil Bhola, University of Michigan, Research aide, 2021

\subsubsection{A list of postdoctoral associates}
  
  (including holders of named
     instructorships and 2- or 3-year terminal assistant professors),
     their Ph.D. institutions, postdoctoral mentors, and post-appointment
     placements.

     I know those from 2017. Can you check and keep it as a record?

     1) From Duan's group
     
     Jessica Xia 
     
     
     2) From Chun’s group:

     Stefan Metzger: 8/2017 — 8/2018, FAU ( Friedrich-Alexander
     University, Erlangen-Nurnberg), FAU
     Qing Cheng: 1/2019 — 1/2021, Xiamen University, Purdue University
     Yiwei Wang: 8/2018 — 07/2022, University of California, Riverside
     Yubin Lu: 08/2022 - now, Huazhong University of Science and Technology

     2) From Igor’s group

     Hyun-Jung Kim, 8/2018  — 8/2020, USC, UCSB

     3) From Sonja’s group

HOW FAR IN THE PAST ARE THESE SUPPOSED TO GO?? 
     Sara Jamshidi: 8/2018 — 8/2020, Penn State, Lake Forest College

     4) From Shuwang’s group

     Pedro Anjos: 8/2019 — present, Federal University of Pernambuco, IIT.




\subsection{Quantitative Demographic Data}
 


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
